---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PMXFrem

The goal of `PMXFrem` is to provide comprehensive post-processing support for Full Random Effects Models (FREM) built in NONMEM. It facilitates the automated conversion of FREM models to Full Fixed Effects Models (FFEM), enabling standard project workflows including Goodness-of-Fit (GOF) diagnostics, Visual Predictive Checks (VPC), and forest-plot based results communication.

## Installation

You can install the development version of `PMXFrem` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pharmetheus/PMXFrem")
```

The vignette sources are in `vignettes/`. Open the `.Rmd` files in RStudio and click Knit to compile, or run `browseVignettes("PMXFrem")` to read the versions built at install time.

To get the latest stable release (including vignettes), please click the appropriate link on the right side of the page. Download the file with a name that starts with PMXFrem. Install `PMXFrem` with the `install.packages` command, e.g.:

```{r, eval = FALSE}
install.packages("path_to_release.tar.gz", repos = NULL, type = "source")
```

If your organizational SOPs require installation via an internal package manager (e.g., Posit Package Manager) or a localized server, please prioritize those methods over GitHub to ensure environment stability and reproducibility.

## Example of using PMXFrem

This is a basic example which shows what a converged FREM model gives a project: an FFEM model to run the usual diagnostics with, a parameter table, and the share of the parameter variability the covariates explain.

First attach the PMXFrem library:
```{r example}
library(PMXFrem)
```

and set the ggplot theme
```{r}
ggplot2::theme_set(ggplot2::theme_bw())
```

### Point at the model and the data

Everything below uses the `run31` example bundled with the package, so it runs
anywhere `PMXFrem` is installed. `run31` is a FREM model with 18 covariates,
built from the base model `run30`.

```{r}
modDevDir <- system.file("extdata", "SimNeb", package = "PMXFrem")
modFile <- file.path(modDevDir, "run31.mod")
extFile <- file.path(modDevDir, "run31.ext")
dataFile <- file.path(modDevDir, "DAT-2-MI-PMX-2-onlyTYPE2-new.csv")
```

A FREM model is described by a handful of structural integers: how many thetas
are structural rather than covariate means, how many etas precede the FREM
block, and so on. `fremModelInfo()` reads them from the model and its `.ext`, so
they do not have to be typed in and kept up to date as the model changes.

```{r}
info <- fremModelInfo(modFile, extFile)
unlist(info[c("numNonFREMThetas", "numFREMThetas", "numParCov", "numSkipOm")])
```

The functions below locate the FREM model through `runno` / `modName` and
`modDevDir` and derive the same integers. Supply them explicitly only to
override the derived values.

### Convert the FREM model to its FFEM counterpart

`createFFEMmodel()` writes the FFEM control stream, and the data set that goes
with it, from the FREM model: the covariate coefficients are computed from the
FREM omega matrix and written into the model, so GOF diagnostics and VPCs can be
run in the ordinary way.

```{r}
td <- tempdir()

ffemMod <- createFFEMmodel(
  runno = 31,
  baserunno = 30,
  modDevDir = modDevDir,
  dataFile = dataFile,
  parNames = c("CL", "V", "MAT"),
  newDataFile = file.path(td, "ffemData31.csv"),
  ffemModName = file.path(td, "run31_ffem.mod"),
  quiet = TRUE
)

head(ffemMod, 8)
```

### Report the parameter estimates

`fremParameterTable()` collects the structural thetas, the covariate-adjusted
variances and the residual error, with uncertainty from sampled parameter
vectors.

```{r}
tab <- fremParameterTable(
  runno = 31,
  modDevDir = modDevDir,
  thetaNum = 2:7,
  omegaNum = c(1, 3, 4, 5),
  sigmaNum = 1,
  quiet = TRUE
)

tab$parameterTable
```

### Describe how the parameters are related to the covariates

The parameter function turns the estimates, one row of covariate values and a
set of etas into the quantities to plot. `createFREMParamFunction()` writes it by
transliterating the FREM model's own `$PK`, so the relationship is not retyped in
R and cannot drift away from the model. The result is source text; nothing is
evaluated until you do it yourself.

```{r}
gen <- createFREMParamFunction(modFile, parameters = c("CL", "V", "MAT"))
```

`verifyFREMParamFunction()` is the check to run before trusting it. It compares
the generated function with the FFEM version of the same model - an independent
transliteration - and confirms that each parameter's covariate coefficient and
eta land where the model says they do.

```{r}
verifyFREMParamFunction(gen, ffemModel = file.path(modDevDir, "run31max1-2.mod"))
```

Read the code against the control stream, then evaluate it:

```{r}
paramFunction <- eval(parse(text = gen$code))
```

### Show how much of the variability the covariates explain

`setupDfCovsEV()` builds one row per covariate, and `getExplainedVar()` computes
the total variability of each parameter together with the part of it each
covariate explains. `type = 0` uses the delta rule, which is the quick way in.

```{r}
dfCovs <- setupDfCovsEV(modFile)

dfres <- getExplainedVar(
  type = 0,
  dfCovs = dfCovs,
  cstrCovariates = c("All", names(dfCovs)),
  functionList = list(paramFunction),
  functionListName = c("CL", "V", "MAT"),
  modDevDir = modDevDir,
  runno = 31,
  quiet = TRUE
)

head(dfres, 4)
```

```{r explained-var, fig.width = 8, fig.height = 6}
plotExplainedVar(dfres, parameters = "CL")
```

## Further information

The above is a simple use case of PMXFrem for a straightforward FREM model.
PMXFrem also assembles FREM data sets, extends an established FREM model with new
structural thetas and IIVs, derives subject-specific variances for missing
covariates, and produces the diagnostics and forest plots a FREM analysis is
reported with. These are described in the package vignettes, which are written to
be read in order:

1. **Quick Start** - a 5-minute copy-pasteable example to verify your installation and run a basic FREM-to-FFEM conversion.
2. **The Walk-Through** - the full workflow, from data assembly to forest plot generation.
3. **The deep dives** - one vignette per topic, for the parts of the workflow that need more than the walk-through gives them.

Once the package is installed you can issue the command: `browseVignettes("PMXFrem")`
