---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PMXFrem
The goal of `PMXFrem` is to provide comprehensive post-processing support for Full Random Effects Models (FREM) built in NONMEM. It facilitates the automated conversion of FREM models to Full Fixed Effects Models (FFEM), enabling standard project workflows including Goodness-of-Fit (GOF) diagnostics, Visual Predictive Checks (VPC), and forest-plot based results communication.

## Installation

### Nominal Instructions 
You can install the development version of `PMXFrem` from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("pharmetheus/PMXFrem")
```

To install a stable release with compiled vignettes, download the latest source package (.tar.gz) from the Releases section and install locally:

```{r,eval=FALSE}
install.packages("path_to_release/PMXFrem_vX.Y.Z.tar.gz", repos = NULL, type = "source")
```

### Organizational Installation

If your organizational SOPs require installation via an internal package manager (e.g., Posit Package Manager) or a localized server, please prioritize those methods over GitHub to ensure environment stability and reproducibility.

## Documentation Roadmap

To get started with `PMXFrem`, we recommend following this documentation sequence:

1. Quick Start: A 5-minute copy-pasteable example to verify your installation and run a basic FREM-to-FFEM conversion.
2. The Walk-Through: A comprehensive tutorial demonstrating the full workflow: from data assembly to forest plot generation.