---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# PMXFrem

The goal of `PMXFrem` is to provide pre- and post-processing
support for Full Random Effects Models (FREM) built in NONMEM. It
facilitates the conversion of FREM models to Full Fixed
Effects Models (FFEM), enabling standard project workflows including
Goodness-of-Fit (GOF) diagnostics, Visual Predictive Checks (VPC), and
forest-plot based results communication.

## Installation

You can install the development version of `PMXFrem` from [GitHub](https://github.com/pharmetheus/PMXFrem) with:

``` r
# install.packages("devtools")
devtools::install_github("pharmetheus/PMXFrem")
```

The vignette sources are in `vignettes/`. Open the `.Rmd` files in RStudio and click Knit to compile, or run `browseVignettes("PMXFrem")` to read the versions built at install time.

To get the latest stable release (including vignettes), download the source package from the [releases page of the GitHub repository](https://github.com/pharmetheus/PMXFrem/releases): the file whose name starts with `PMXFrem` and ends in `.tar.gz`. Install it with the `install.packages` command, e.g.:

```{r, eval = FALSE}
install.packages("path_to_release.tar.gz", repos = NULL, type = "source")
```

If your organizational SOPs require installation via an internal package manager (e.g., Posit Package Manager) or a localized server, please prioritize those methods over GitHub to ensure environment stability and reproducibility.

## Further information

A FREM analysis has more moving parts than a short example can usefully show, so
the documentation takes it a piece at a time. PMXFrem assembles FREM data sets,
converts a converged FREM model to its FFEM counterpart, extends an established
model with new structural thetas and IIVs, derives subject-specific variances
for missing covariates, and produces the parameter tables, diagnostics, forest
plots and explained-variability plots a FREM analysis is reported with. The
vignettes are written to be read in order:

1. **Quick Start** - a 5-minute copy-pasteable example to verify your installation and run a basic FREM-to-FFEM conversion.
2. **The Walk-Through** - the full workflow, from data assembly to forest plot generation.
3. **The deep dives** - one vignette per topic, for the parts of the workflow that need more than the walk-through gives them.

Once the package is installed you can issue the command: `browseVignettes("PMXFrem")`
