---
title: "3. Deep Dive: Explained Variability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Deep Dive: Explained Variability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8,
  warning = FALSE,
  message = FALSE
)
```

## Introduction
This vignette explains how to use the `PMXFrem` package to visualize the degree to which covariates manage to explain variability in the parameters of interest. 

The workflow is divided into five logical steps:

1. **Pre-requisite Setup:** Loading models and datasets.
2. **Covariate Scenarios:** Defining the specific covariate combinations to isolate.
3. **Parameter Function Definition:** Writing the structural equations to evaluate.
4. **Computing Explained Variability:** Deriving the variance metrics using specific integration methods.
5. **Visualization:** Rendering the explained variability bar plots.

## 1. Setting up the Pre-requisite

Load the package and identify which NONMEM FREM run is going to be used.

```{r setup}
library(PMXFrem) 
library(ggplot2)
library(dplyr)
library(data.table)

# Enforce a clean global theme
set.seed(865765)
theme_set(theme_bw(base_size = 14))

# Specify the name of the FREM run and collect file paths
runname <- "run31"
modDevDir <- system.file("extdata/SimNeb", package = "PMXFrem")
fremFiles <- getFileNames(modName = runname, modDevDir = modDevDir)

numNonFREMThetas <- 7
numSkipOm        <- 2

dfData <- fread(system.file("extdata/SimNeb/DAT-2-MI-PMX-2-onlyTYPE2-new.csv", package="PMXFrem"), data.table = FALSE) %>% 
  distinct(ID, .keep_all = TRUE)
```

## 2. Defining the Covariate Scenarios

We use the package's internal `setupdfCovs()` function to dynamically construct the covariate isolation matrix. Because `FOOD` is a time-varying categorical covariate in the base dataset, we manually append it as an `additionalCov` and lock its scenario state to `1` (Fed).

```{r cov_setup}
# Robustly generate the covariate matrix
dfCovs <- setupdfCovs(
  modFileName = fremFiles$mod, 
  additionalCovs = c("FOOD")
)

# Lock the time-varying covariate state
dfCovs$FOOD <- 1

# Define the exact names corresponding to the rows in dfCovs
cstrCovariates <- c("All", names(dfCovs))
```

## 3. Defining the Parameter Function

We must supply a function that mimics the structural relationships in the FFEM model. This function is conceptually identical to the one used for Forest plots, with one critical addition: it must handle the `etas` argument to allow the package to integrate over the structural variance.

**CRITICAL WARNING:** Crucially, even if your original dataset has zero missing values, you must build defensive fallback logic for missing covariates (e.g., `dfrow$FOOD != -99`). Why? Because the `dfCovs` matrix intentionally injects `-99` into rows to mathematically isolate specific covariate effects during the variance calculation.

```{r define_function}
paramFunctionExpVar <- function(basethetas, covthetas, dfrow, etas, ...) {

  # Defensive fallback logic for isolating covariate effects
  FRELFOOD <- 1
  if (any(names(dfrow) == "FOOD") && dfrow$FOOD != -99 && dfrow$FOOD == 0) {
    FRELFOOD <- (1 + basethetas[7])
  }

  MATFOOD <- 1
  if (any(names(dfrow) == "FOOD") && dfrow$FOOD != -99 && dfrow$FOOD == 0) {
    MATFOOD <- (1 + basethetas[6])
  }

  TVFREL <- basethetas[1]
  TVCL   <- basethetas[2]
  TVV    <- basethetas[3]
  TVMAT  <- basethetas[4]
  TVD1   <- basethetas[5]

  MU_2 <- TVD1
  MU_3 <- log(TVFREL)
  MU_4 <- log(TVCL)
  MU_5 <- log(TVV)
  MU_6 <- log(TVMAT)

  D1FR  <- MU_2 + etas[2]
  FREL  <- FRELFOOD * exp(MU_3)
  CL    <- exp(MU_4 + covthetas[1] + etas[3])
  V     <- exp(MU_5 + covthetas[2] + etas[4])
  MAT   <- MATFOOD * exp(MU_6 + covthetas[3] + etas[5])

  D1   <- MAT * (1 - TVD1)
  KA   <- 1 / (MAT - D1)

  # Compute AUC after an 80 mg dose
  AUC <- 80 / (CL / FREL)

  return(list(CL, FREL, AUC, V, MAT))
}

functionListName <- c("CL", "Frel", "AUC", "V", "MAT")
```

## 4. Computing Explained Variability

The `getExplainedVar()` function computes the total and explained variability. [cite_start]The variability can be related to the total variability in the parameter ($TOTVAR$) or the total covariate variability in the parameter ($TOTCOVVAR$)[cite: 1341, 1342, 1343].

The `dfres` output contains three critical metrics:
* [cite_start]$TOTVAR$: The total variance, calculated using the Delta method (Type 0), EBEs (Type 1), or $\eta$ samples (Type 2, 3)[cite: 1346, 1347, 1348, 1349].
* [cite_start]$TOTCOVVAR$: Calculated using all covariates with $\eta=0$, across all observed covariate values[cite: 1358, 1359, 1360].
* [cite_start]$COVVAR$: Calculated with one (or a combination) of covariates, setting $\eta=0$[cite: 1374, 1375].

### Calculation Methods (`type`)
[cite_start]The methodology is dictated by the `type` argument, and each has distinct mathematical properties[cite: 1397]:
* **Type 0 (Delta Method):** Based on omega calculations. [cite_start]It is fast and guarantees that the explained variability of all covariates will sum to exactly 100%[cite: 1398].
* **Type 1 (EBEs):** Uses Empirical Bayes Estimates. [cite_start]It is subject to $\eta$ shrinkage but acknowledges individual information richness[cite: 1399].
* **Type 2 ($\eta$ Samples):** Uses $\eta$ samples and integrates over observed FFEM covariates. [cite_start]Useful if the model contains non-FREM covariates[cite: 1400].
* [cite_start]**Type 3 ($\eta$ Samples Optimized):** Similar to Type 2 but does not acknowledge FFEM covariates, making it significantly faster[cite: 1401].

[cite_start]*Note: With Types 1, 2, and 3, it is theoretically possible for a single covariate to explain more than all covariates combined (i.e., > 100%) due to parameter correlations and structural non-linearities[cite: 1402].*

```{r compute_var}
dfres <- getExplainedVar(
  type             = 2,
  data             = dfData,
  dfCovs           = dfCovs,
  numNonFREMThetas = numNonFREMThetas,
  numSkipOm        = numSkipOm,
  functionList     = list(paramFunctionExpVar),
  functionListName = functionListName,
  cstrCovariates   = cstrCovariates,
  modDevDir        = modDevDir,
  modName          = runname,
  ncores           = 1,         # Required sequential execution for CRAN
  numETASamples    = 100,
  quiet            = TRUE,
  seed             = 123
)
```

## 5. Visualization

We plot the results using `plotExplainedVar()`. To ensure reproducibility and prevent silent misalignment, we map our labels using a programmatic dictionary.

```{r plot_var}
# 1. Define robust label mappings
label_map <- c(
  "All" = "All Covariates", "AGE" = "Age", "ALT" = "ALT", "AST" = "AST", 
  "BILI" = "Bilirubin", "BMI" = "BMI", "BSA" = "BSA", "CRCL" = "Creatinine Clearance", 
  "ETHN" = "Ethnicity", "GENO2" = "Genotype", "HT" = "Height", "LBM" = "Lean Body Mass", 
  "NCIL" = "NCI Grade", "RACEL" = "Race", "SEX" = "Sex", "SMOK" = "Smoking Status", 
  "WT" = "Weight", "FOOD" = "Food Status"
)

# 2. Safely translate cstrCovariates into clean labels
mapped_labels <- unname(sapply(cstrCovariates, function(x) {
  ifelse(x %in% names(label_map), label_map[[x]], x)
}))

# 3. Plot Total Explained Variability (maxVar = 1)
# Visualizes the fraction: 100 * (COVVAR / TOTVAR)
plotExplainedVar(
  dfres, 
  parameters = "AUC", 
  maxVar = 1, 
  covariateLabels = mapped_labels
) + labs(title = "Explained Part of Total Variability (AUC)")

# 4. Plot Proportion of Explainable Variability (maxVar = 2)
# Visualizes the fraction: 100 * (COVVAR / TOTCOVVAR)
plotExplainedVar(
  dfres, 
  parameters = "AUC", 
  maxVar = 2, 
  covariateLabels = mapped_labels
) + labs(title = "Explained Part of Explainable Variability (AUC)")
```