---
title: "3. Deep Dive: Forest Plots"

output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Forest Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 8,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette explains how to use the `PMXFrem` and `PMXForest` packages to create a Forest plot based on a FREM run, incorporating its associated uncertainty. 

The uncertainty may stem from a NONMEM covariance step (a `.cov` file), a PsN bootstrap/SIR run (a `raw_results` file), or a custom data frame of parameter vectors. 

The workflow is divided into four stages:

1. **Pre-requisite Setup:** Loading models, estimates, and parameter samples.
2. **Covariate Scenarios:** Defining the specific covariate combinations to visualize.
3. **Data Generation:** Computing the relative parameter changes and their confidence intervals.
4. **Visualization:** Rendering the Forest plot with programmatic, robust labeling.

---

## 1. Pre-requisite Setup

First, we load the required packages and extract the base file paths, parameter estimates, and covariate names from our converged FREM model.

```{r setup}
library(PMXFrem)
library(PMXForest)
library(ggplot2)
library(dplyr)
library(data.table)

# Enforce a clean global theme
theme_set(theme_bw(base_size = 14))
set.seed(865765) 

# Identify the FREM run and collect file paths
runname <- "run31"
modDevDir <- system.file("extdata/SimNeb", package = "PMXFrem")
fremFiles <- getFileNames(modName = runname, modDevDir = modDevDir)

# Extract final parameter estimates and annotated covariate names
dfExt <- subset(getExt(extFile = fremFiles$ext), ITERATION == "-1000000000") 
covNames <- getCovNames(modFile = fremFiles$mod)                             
```

Next, we extract the uncertainty distribution. In this example, we draw 175 multivariate normal samples using the variance-covariance matrix from the NONMEM `.cov` file. 

```{r get_samples}
# Get 175 samples from the .cov file
dfParameters <- getSamples(fremFiles$cov, extFile = fremFiles$ext, n = 175)

# Enforce the first row as the final deterministic estimates
dfParameters[1, ] <- dfExt[1, -1]
```

### The Parameter Function
We must define a list of functions specifying how the parameters are calculated. Because we are plotting the *typical* parameter value for a given covariate, these functions only utilize the fixed effects. 

Note that the functional form for the covariate-parameter relationships depends on the parameter functional form (usually log-normal):
$$CL = \theta_{CL} \cdot \exp(\eta_{CL})$$

As well as the covariate form in the FREM model (usually normal):
$$COV = \theta_{COV} + \eta_{COV}$$

```{r define_functions}
# The parameter function list for log-normal parameters
functionList <- list(
  function(basethetas, covthetas, ...) { return(basethetas[1] * exp(covthetas[1])) }, 
  function(basethetas, covthetas, ...) { return(basethetas[2] * exp(covthetas[2])) }
)

functionListName <- c("CL", "V1")
```

---

## 2. Defining Covariate Scenarios

We need to define the specific covariate combinations (univariable or multivariable) to plot. A value of `-99` or `NA` indicates that the covariate is at the reference (population typical) value. 

While you can build this matrix manually, `PMXForest::getCovStats()` is the recommended approach. It automatically scans your clinical dataset and calculates the 5th and 95th percentiles (or your chosen quantiles) for continuous covariates, and identifies categorical levels.

```{r define_scenarios}
# Load clinical data
data <- fread(system.file("extdata/SimNeb/DAT-2-MI-PMX-2-onlyTYPE2-new.csv", package="PMXFrem"))

# Generate covariate matrix dynamically using the 5th and 90th percentiles
dfCovs <- createInputForestData(
  getCovStats(data, covNames$orgCovNames, probs = c(0.05, 0.90), missVal = -99),
  iMiss = -99
)

head(dfCovs, 5)
```

---

## 3. Generate Forest Plot Data

With our scenarios defined, we pass everything to `getForestDFFREM()`. This function computes the relative parameter changes across all bootstrapped/sampled parameter vectors to construct the confidence intervals.

```{r compute_forest}
dfres <- getForestDFFREM(
  dfCovs           = dfCovs,
  numNonFREMThetas = 7,
  numSkipOm        = 2,
  dfParameters     = dfParameters,
  covNames         = covNames$covNames,
  functionList     = functionList,
  functionListName = functionListName,
  quiet            = TRUE
)
```

---

## 4. Visualization & Labeling

Hardcoding dozens of row labels is dangerous and non-reproducible. Instead, we use `PMXFrem::generateCovNames()`. This utility function parses the `dfCovs` scenarios and applies user-defined unit mappings and categorical translations dynamically.

```{r plot_forest, fig.height=14, fig.width=16, out.width="100%"}
# 1. Define translation maps
unit_map <- c(
  AGE = "years", WT = "kg", BMI = "kg/m2", BSA = "m2", 
  CRCL = "mL/min", BILI = "mg/dL", AST = "IU", ALT = "IU", 
  HT = "cm", LBM = "kg"
)

label_map <- c(
  "SEX=0" = "Male", "SEX=1" = "Female", 
  "RACEL=1" = "Caucasian", "RACEL=2" = "Black", "RACEL=3" = "Other", "RACEL=4" = "Hispanic",
  "SMOK=0" = "Non-smoking", "SMOK=1" = "Smoking", 
  "NCIL=0" = "NCI Grade 0", "NCIL=1" = "NCI Grade 1", "NCIL=2" = "NCI Grade 2", 
  "GENO2=0" = "Poor Metabolizer", "GENO2=1" = "Extensive Metabolizer"
)

group_map <- c(
  "AGE" = "Age", "ALT" = "ALT", "AST" = "AST", "BILI" = "Total Bilirubin", 
  "BMI" = "BMI", "BSA" = "BSA", "CRCL" = "Creatinine Clearance", 
  "ETHN" = "Ethnicity", "GENO2" = "Genotype", "HT" = "Height", 
  "LBM" = "Lean Body Weight", "NCIL" = "NCI", "RACEL" = "Race", 
  "SEX" = "Sex", "SMOK" = "Smoking Status", "WT" = "Weight"
)

# 2. Programmatically generate and map robust labels and groups directly to the results
dfres_formatted <- generateCovNames(
  dfCovs, 
  dfres = dfres, 
  unit_map = unit_map, 
  label_map = label_map,
  group_map = group_map
)

# 3. Render Plot
PMXForest::forestPlot(
  dfres_formatted,
  tabTextSize = 14,
  strip_right_size = 11
)
```