---
title: "3. Deep Dive: Modifying FREM Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Modifying FREM Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

The Full Random Effects Model (FREM) relies on a highly interdependent architecture between the NONMEM control stream and the underlying dataset. Manually editing these files to add or remove covariates can be time-consuming and introduces a risk for human error into the pharmacometric workflow.

The `PMXFrem` package provides the `updateFREMmodel()` function to programmatically manage these complex structural updates. 

This deep dive demonstrates four primary capabilities:

* Removing covariates from an existing model.
* Adding new covariates to a model.
* Appending new individuals to an existing FREM dataset.
* Updating a model's initial parameter estimates directly from a converged `.ext` file.

## 1. Pre-requisite Setup

First, we load the package. To ensure system reproducibility, we map all base files from the package's `extdata` directory and explicitly route all generated artifacts into a secure temporary directory (`tempdir()`).

```{r setup}
library(PMXFrem)
library(dplyr)
library(data.table)

# Isolate all generated outputs
td <- tempdir()

# Define paths to validated example files
frem_model_path <- system.file("extdata/SimNeb/run31.mod", package = "PMXFrem")
frem_ext_path   <- system.file("extdata/SimNeb/run31.ext", package = "PMXFrem")
frem_data_path  <- system.file("extdata/SimNeb/frem_dataset.dta", package = "PMXFrem")
ffem_data_path  <- system.file("extdata/SimNeb/DAT-2-MI-PMX-2-onlyTYPE2-new.csv", package = "PMXFrem")

# Copy base model and ext files into the temporary directory for safe manipulation
file.copy(frem_model_path, td)
file.copy(frem_ext_path, td)
model_in_td <- file.path(td, "run31.mod")
```

---

## 2. Use Case 1: Removing Covariates

You may sometimes need to remove covariates from an existing model. Here, we remove `WT`, `SEX`, and `AGE` from `run31.mod`.

The function requires the base model's structural configuration: `numNonFREMThetas` (the number of structural fixed effects) and `numSkipOm` (the number of standard variances, such as IIV or RUV, that precede the FREM `$OMEGA BLOCK`).

```{r remove_covariates}
result_remove <- updateFREMmodel(
  strFREMModel      = model_in_td,
  strFREMData       = frem_data_path,
  strFFEMData       = ffem_data_path,
  cstrRemoveCov     = c("SEX", "WT", "AGE"),
  strNewFREMData    = file.path(td, "frem_data_simplified.csv"),
  numNonFREMThetas  = 7,
  numSkipOm         = 2,
  bWriteData        = TRUE,   # Generate the updated dataset
  bWriteMod         = TRUE,   # Generate the updated control stream
  quiet             = TRUE,
  cstrKeepCols      = c("ID", "TIME", "AMT", "EVID", "RATE", "DV", "FOOD", "FREMTYPE")
)

new_model_path_remove <- file.path(td, "run31_new.mod")
print(paste("Simplified model successfully written to:", new_model_path_remove))
```

**Architectural Operations:**

* The algorithm drops dataset rows associated with the removed covariates and dynamically re-indexes the remaining `FREMTYPE` identifiers to prevent gaps in the NONMEM `$ERROR` block.
* The covariance matrix (`$OMEGA`) and typical values (`$THETA`) are systematically adjusted.

---

## 3. Use Case 2: Adding Covariates

Next, we take the simplified model generated above and insert a new covariate (`AGE`) back into the architecture. We explicitly set `bRecodeDichotomous = TRUE` as a standard practice to ensure any 2-level categorical covariates are safely zero-indexed.

```{r add_covariates}
# Utilize the newly generated model and dataset as our baseline
model_to_add_to <- new_model_path_remove
data_to_add_to  <- file.path(td, "frem_data_simplified.csv")

# The function requires the corresponding .ext file for the input model
file.copy(file.path(td, "run31.ext"), file.path(td, "run31_new.ext"))

result_add <- updateFREMmodel(
  strFREMModel       = model_to_add_to,
  strFREMData        = data_to_add_to,
  strFFEMData        = ffem_data_path,
  cstrContCovsToAdd  = "AGE", 
  strNewFREMData     = file.path(td, "frem_data_with_age.csv"),
  numNonFREMThetas   = 7,
  numSkipOm          = 2,
  bWriteData         = TRUE,
  bWriteMod          = TRUE,
  bRecodeDichotomous = TRUE, # Safely zero-indexes 2-level categorical factors
  quiet              = TRUE,
  cstrKeepCols       = c("ID", "TIME", "AMT", "EVID", "RATE", "DV", "FOOD", "FREMTYPE")
)

print(paste("Expanded model successfully written to:", file.path(td, "run31_new_new.mod")))
```

**Architectural Operations:**

* The base clinical data (`strFFEMData`) is scanned to extract the new covariate's mean and variance, which seed the new `$THETA` and `$OMEGA` parameters.
* Dummy observation rows (`EVID=2`) are appended to the dataset for the new covariate under a newly assigned `FREMTYPE`.

---

## 4. Use Case 3: Adding New Individuals

An advantage of the FREM architecture is the ability to process new subjects through a converged model without altering the matrix structure. Here, we simulate 10 new individuals and fold them into the dataset.

```{r add_individuals}
# 1. Load the pristine FREM and FFEM datasets
frem_df <- fread(frem_data_path, data.table = FALSE) %>% select(-ODV)
ffem_df <- fread(ffem_data_path, data.table = FALSE)

# 2. Simulate 10 new individuals by shifting IDs
ffem_new_ids <- ffem_df[ffem_df$ID <= 10, ] 
ffem_new_ids$ID <- ffem_new_ids$ID + max(frem_df$ID)

# 3. Append the individuals to the dataset
result_new_ids <- updateFREMmodel(
  strFREMModel      = model_in_td,
  strFREMData       = frem_df,        # Pass dataset directly via memory
  strFFEMData       = ffem_new_ids,   # Include the new IDs
  strNewFREMData    = file.path(td, "frem_data_appended.csv"),
  numNonFREMThetas  = 7,
  numSkipOm         = 2,
  bWriteData        = TRUE, 
  bWriteMod         = FALSE,          # The NONMEM model logic does not change
  quiet             = TRUE,
  cstrKeepCols      = c("ID", "TIME", "AMT", "DV", "EVID", "RATE", "FREMTYPE"),
  cstrSetToZero     = c("AMT", "EVID", "RATE")
)

print(paste("Original Max ID:", max(frem_df$ID), "| New Max ID:", max(result_new_ids$data$ID)))
```

**Architectural Operations:**

* The algorithm expands the dataset horizontally, mapping the structural and covariate dummy rows for the new IDs while keeping the control stream untouched (`bWriteMod = FALSE`).

---

## 5. Use Case 4: Updating Initial Estimates

Due to the size of typical FREM models, it is not feasible to update initial estimates manually. The PsN tool `update_inits` can be used for this, or you can use `updateFREMmodel` to do it by specifying a "NoData" update.

```{r update_inits}
result_inits <- updateFREMmodel(
  strFREMModel       = model_in_td,
  strUpdateType      = "NoData",     # Skip dataset processing entirely
  basenames_th       = c("CL", "V", "MAT", "D1", "FRELFOOD", "MATFOOD"),
  basenames_om       = c("RUV", "D1", "CL", "V", "MAT"),
  numNonFREMThetas   = 7,
  numSkipOm          = 2,
  bWriteData         = FALSE,
  bWriteMod          = TRUE,
  bWriteFIX          = TRUE,         # Explicitly carry over FIX flags
  quiet              = TRUE
)

print("Control stream initial estimates successfully updated from the .ext file.")
```

**Architectural Operations:**

* `updateFREMmodel()` parses the `-1000000000` iteration row from the provided `.ext` file.
* The `$THETA`, `$OMEGA`, and `$SIGMA` blocks in the `.mod` file are surgically replaced, and user-defined `basenames` are appended as clean comments for maximum readability.