---
title: "3. Deep Dive: Explained Variability"
author: "PMXFrem Development Team"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Explained Variability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8,
  warning = FALSE,
  message = FALSE
)
```

# Introduction
This vignette explains how to use the `PMXFrem` package to visualize the degree to which covariates manage to explain variability in the parameters of interest. 

The workflow is divided into five logical steps:

1. **Pre-requisite Setup:** Loading models and datasets.
2. **Covariate Scenarios:** Defining the specific covariate combinations to isolate.
3. **Parameter Function Definition:** Writing the structural equations to evaluate.
4. **Computing Explained Variability:** Deriving the variance metrics using specific integration methods.
5. **Visualization:** Rendering the explained variability bar plots.

# 1. Setting up the Pre-requisite

Load the package and identify which NONMEM FREM run is going to be used.

```{r setup}
library(PMXFrem)
library(ggplot2)
library(dplyr)
library(data.table)

# One theme for every figure below
set.seed(865765)
theme_set(theme_bw(base_size = 14))

# Specify the name of the FREM run and collect file paths
runname <- "run31"
modDevDir <- system.file("extdata/SimNeb", package = "PMXFrem")
fremFiles <- getFileNames(modName = runname, modDevDir = modDevDir)

dfData <- fread(system.file("extdata/SimNeb/DAT-2-MI-PMX-2-onlyTYPE2-new.csv", package = "PMXFrem"), data.table = FALSE) %>%
  distinct(ID, .keep_all = TRUE)
```

# 2. Defining the Covariate Scenarios

`setupDfCovsEV()` builds the covariate matrix from the model, one row per covariate. 

`dfCovs` is read differently here than in a Forest plot. In a Forest plot, `dfCovs` dictates the exact numerical covariate values to visualize (e.g., `WT = 115`). For explained variability, `dfCovs` acts purely as an indicator matrix **for the FREM covariates**. For these covariates, the actual numerical value assigned does not impact the calculation; setting it to anything other than the missing covariate indicator (typically `-99`) simply flags *which* covariates should be actively evaluated for that specific scenario row.

**FFEM covariates work differently.** A covariate outside the FREM matrix, such as `FOOD`, is read by the parameter function itself from `dfrow`, and where that value comes from depends on `type`:

* **Types 1, 2 and 3** use each subject's own value from `data`. The value in `dfCovs` is not used.
* **Type 0** uses the first row of `dfCovs` for every calculation.

In a covariate row, `dfrow` holds only the covariates that are not `-99` in that row. A covariate passed as `conditionalCovs` is set to `1` in *every* row, which is what makes it conditional: each FREM covariate's row then reports what that covariate explains with `FOOD` in the model. Its own contribution is part of every row, so those rows are not comparable with the `FOOD` row. The FREM covariates themselves are still isolated one row at a time, and the function must still fall back to the baseline effect when a column is missing (Section 3).

```{r cov_setup}
# Build the covariate matrix; FOOD is on in every row
dfCovs <- setupDfCovsEV(
  modFileName = fremFiles$mod,
  conditionalCovs = c("FOOD")
)

# Define the exact names corresponding to the rows in dfCovs
cstrCovariates <- c("All", names(dfCovs))
```

# 3. The Parameter Function

`getExplainedVar()` needs a function that returns the parameters to decompose,
given the model's thetas, the FREM covariate coefficients and a set of etas.
Unlike the Forest plot case, it **must** take `etas`: the whole point is to
integrate over the structural variability.

`createFREMParamFunction()` writes that function from the model's own `$PK`, so
the algebra is not retyped here and cannot drift away from the control stream:

```{r define_function}
gen <- createFREMParamFunction(
  fremModel  = fremFiles$mod,
  parameters = c("CL", "FREL", "V", "MAT"),
  secondary  = list(AUC = "80 / (CL / FREL)"),
  quiet      = TRUE
)

paramFunction <- eval(parse(text = gen$code))
functionListName <- gen$functionListName
functionListName
```

`AUC` is a secondary parameter. The 80 mg dose it is taken over is a property of
the question being asked rather than of `$PK`, so it is supplied at the call
site; the Secondary Parameters deep dive covers that in full.

The result is source text, and worth reading against the control stream before
it is used. `verifyFREMParamFunction()` is the mechanical half of that check: it
compares the generated function with the FFEM version of the same model and
confirms that each parameter's covariate coefficient and eta land where the
model says they do.

```{r verify_function, comment = ""}
verifyFREMParamFunction(gen, ffemModel = file.path(modDevDir, "run31max1-2.mod"))
```

**FFEM covariates are handled for you.** `FOOD` is part of this model's `$PK`
but not of its FREM covariate matrix, so the generated function carries it -
including what to do when `dfrow` has no `FOOD` column, which is the case in
every covariate row that is not its own. The reference comes from the control
stream, here the branch PsN marks `; Most common`. Written by hand, that
fallback is yours to get right: a function that reads `dfrow$FOOD` where the
column is absent gets `NULL`, and `getExplainedVar()` stops rather than
silently dropping the parameter.

# 4. Computing Explained Variability

The `getExplainedVar()` function computes the total and explained variability. The variability can be related to the total variability in the parameter ($TOTVAR$) or the total covariate variability in the parameter ($TOTCOVVAR$).

The `dfres` output contains three main metrics:

* $TOTVAR$: The total variance, calculated using the Delta method (Type 0), EBEs (Type 1), or $\eta$ samples (Type 2, 3).
* $TOTCOVVAR$: Calculated using all covariates with $\eta=0$, across all observed covariate values.
* $COVVAR$: Calculated with one (or a combination) of covariates, setting $\eta=0$.

## Calculation Methods (`type`)
The methodology is dictated by the `type` argument, and each has distinct mathematical properties:

* **Type 0 (Delta Method):** Based on omega calculations. It is fast, but a first-order approximation.
* **Type 1 (EBEs):** Uses Empirical Bayes Estimates. It is subject to $\eta$ shrinkage but acknowledges individual information richness.
* **Type 2 ($\eta$ Samples):** Uses $\eta$ samples and integrates over observed FFEM covariates. Useful if the model contains non-FREM covariates.
* **Type 3 ($\eta$ Samples Optimized):** Similar to Type 2 but does not acknowledge FFEM covariates, making it significantly faster.

*Note: With every type, the variability explained by the single covariates need not add up to the variability explained by all of them together: correlated covariates share explained variability. With Types 1, 2, and 3, a single covariate can even explain more than all covariates combined (i.e., > 100%) due to parameter correlations and structural non-linearities.*

```{r compute_var}
dfres <- getExplainedVar(
  type             = 2,
  data             = dfData,
  dfCovs           = dfCovs,
  functionList     = list(paramFunction),
  functionListName = functionListName,
  cstrCovariates   = cstrCovariates,
  modDevDir        = modDevDir,
  modName          = runname,
  ncores           = 1,
  numETASamples    = 100,
  quiet            = TRUE,
  seed             = 123
)
```

`numNonFREMThetas` and `numSkipOm` are not passed: when the FREM model is
located through `modName` / `modDevDir` (or `runno`), `getExplainedVar()`
derives them -- and `covNames` -- from the model and its `.ext` via
`fremModelInfo()`. Supply them explicitly only to override the derived values.

# 5. Visualization

We plot the results using `plotExplainedVar()`. To ensure reproducibility and prevent silent misalignment, we map our labels using a programmatic dictionary.

```{r plot_var}
# 1. Map the covariate names to the labels for the plot
label_map <- c(
  "All" = "All Covariates", "AGE" = "Age", "ALT" = "ALT", "AST" = "AST",
  "BILI" = "Bilirubin", "BMI" = "BMI", "BSA" = "BSA", "CRCL" = "Creatinine Clearance",
  "ETHNIC" = "Ethnicity", "GENO2" = "Genotype", "HT" = "Height", "LBWT" = "Lean Body Weight",
  "NCIL" = "NCI Grade", "RACEL" = "Race", "SEX" = "Sex", "SMOK" = "Smoking Status",
  "WT" = "Weight", "FOOD" = "Food Status"
)

# 2. Safely translate cstrCovariates into clean labels
mapped_labels <- unname(sapply(cstrCovariates, function(x) {
  ifelse(x %in% names(label_map), label_map[[x]], x)
}))

# 3. Plot Total Explained Variability (maxVar = 1)
# Visualizes the fraction: 100 * (COVVAR / TOTVAR)
plotExplainedVar(
  dfres,
  parameters = "AUC",
  maxVar = 1,
  covariateLabels = mapped_labels
) + labs(title = "Explained Part of Total Variability (AUC)")

# 4. Plot Proportion of Explainable Variability (maxVar = 2)
# Visualizes the fraction: 100 * (COVVAR / TOTCOVVAR)
plotExplainedVar(
  dfres,
  parameters = "AUC",
  maxVar = 2,
  covariateLabels = mapped_labels
) + labs(title = "Explained Part of Explainable Variability (AUC)")
```
