---
title: "3. Deep Dive: Forest Plots"

output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Forest Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 8,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

This vignette explains how to use the `PMXFrem` and `PMXForest` packages to create a Forest plot based on a FREM run, incorporating its associated uncertainty. 

The uncertainty may stem from a NONMEM covariance step (a `.cov` file), a PsN bootstrap/SIR run (a `raw_results` file), or a custom data frame of parameter vectors. 

The workflow is divided into four stages:

1. **Pre-requisite Setup:** Loading models, estimates, and parameter samples.
2. **Covariate Scenarios:** Defining the specific covariate combinations to visualize.
3. **Data Generation:** Computing the relative parameter changes and their confidence intervals.
4. **Visualization:** Drawing the plot, with the row labels generated rather than typed.

---

# 1. Pre-requisite Setup

First, we load the required packages and extract the base file paths, parameter estimates, and covariate names from our converged FREM model.

```{r setup}
library(PMXFrem)
library(PMXForest)
library(ggplot2)
library(dplyr)
library(data.table)

# One theme for every figure below
theme_set(theme_bw(base_size = 14))
set.seed(865765)

# Identify the FREM run and collect file paths
runname <- "run31"
modDevDir <- system.file("extdata/SimNeb", package = "PMXFrem")
fremFiles <- getFileNames(modName = runname, modDevDir = modDevDir)

# Extract final parameter estimates and annotated covariate names
dfExt <- subset(getExt(extFile = fremFiles$ext), ITERATION == "-1000000000")
covNames <- getCovNames(modFile = fremFiles$mod)
```

Next, we extract the uncertainty distribution. In this example, we draw 175 multivariate normal samples using the variance-covariance matrix from the NONMEM `.cov` file. 

```{r get_samples}
# Get 175 samples from the .cov file
dfParameters <- getSamples(fremFiles$cov, extFile = fremFiles$ext, n = 175)

# Row 1 is the final estimates, so the plot has a deterministic reference
dfParameters[1, ] <- dfExt[1, -1]
```

## The Parameter Function
We must define a list of functions specifying how the parameters are calculated. Because we are plotting the *typical* parameter value for a given covariate, these functions use the fixed effects only. 

Note that the functional form for the covariate-parameter relationships depends on the parameter functional form (usually log-normal):
$$CL = \theta_{CL} \cdot \exp(\eta_{CL})$$

As well as the covariate form in the FREM model (usually normal):
$$COV = \theta_{COV} + \eta_{COV}$$

```{r define_functions}
# The parameter function list for log-normal parameters. The indices are the
# model's own: in run31 THETA(1) is TVFREL, so CL is basethetas[2], and CL is
# the model's first FREM parameter, so its covariate effect is covthetas[1].
functionList <- list(
  function(basethetas, covthetas, ...) {
    return(basethetas[2] * exp(covthetas[1]))
  },
  function(basethetas, covthetas, ...) {
    return(basethetas[3] * exp(covthetas[2]))
  }
)

functionListName <- c("CL", "V")
```

Getting those two index sets right is the fiddly part of writing a parameter
function, and nothing downstream can tell you when they are wrong: a Forest plot
shows ratios to a reference, so a structural theta taken from the wrong index
cancels out and the plot looks entirely reasonable.

`createFREMParamFunction()` removes that risk by transliterating the model's own
`$PK` instead, and one generated function serves both plot types -- call it with
`etas = 0` for a Forest plot, with sampled etas for an explained-variability
plot. It is not used here because this deep dive is about the *inputs* to a
Forest plot, and a two-line function keeps the contract those inputs have to
satisfy visible. The Secondary Parameters deep dive generates the function and
draws the Forest plot from it; the Explained Variability deep dive does the
same for the other plot type.

---

# 2. Defining Covariate Scenarios

We need to define the specific covariate combinations (univariable or multivariable) to plot. A value of `-99` or `NA` indicates that the covariate is at the reference (population typical) value. 

While you can build this matrix manually, `PMXForest::getCovStats()` is the recommended approach. It automatically scans your clinical dataset and calculates the 5th and 95th percentiles (or your chosen quantiles) for continuous covariates, and identifies categorical levels.

```{r define_scenarios}
# Load clinical data
data <- fread(system.file("extdata/SimNeb/DAT-2-MI-PMX-2-onlyTYPE2-new.csv", package = "PMXFrem"))

# Build the covariate matrix from the data, at the 5th and 90th percentiles
dfCovs <- createInputForestData(
  getCovStats(data, covNames$orgCovNames, probs = c(0.05, 0.90), missVal = -99),
  iMiss = -99
)

head(dfCovs, 5)
```

---

# 3. Generate Forest Plot Data

With our scenarios defined, we pass everything to `getForestDFFREM()`. This function computes the relative parameter changes across all bootstrapped/sampled parameter vectors to construct the confidence intervals.

We point it at the FREM model with `modName` / `modDevDir` (or `runno`), so
`numNonFREMThetas`, `numSkipOm` and `covNames` are derived from the model and
its `.ext` (via `fremModelInfo()`) rather than passed by hand. Supply them
explicitly only to override the derived values.

```{r compute_forest}
dfres <- getForestDFFREM(
  dfCovs           = dfCovs,
  modName          = runname,
  modDevDir        = modDevDir,
  dfParameters     = dfParameters,
  functionList     = functionList,
  functionListName = functionListName,
  quiet            = TRUE
)
```

---

# 4. Visualization & Labeling

Typing dozens of row labels out by hand means they drift away from `dfCovs` the first time a covariate is added or dropped. `generateCovNames()` reads the scenarios and applies the unit and level mappings given below, so the labels follow the data.

It is meant to be used in two steps. Run it first with `print_template = TRUE`:
it prints an empty mapping skeleton listing every covariate group, and every
covariate level, that this `dfCovs` actually contains.

<details>
<summary>The mapping template for this `dfCovs` (click to expand)</summary>

```{r label_template, comment = ""}
generateCovNames(dfCovs, print_template = TRUE)
```

</details>

That skeleton is worth starting from rather than writing the maps from memory.
A map written by hand goes stale quietly: a key for a covariate that is not in
`dfCovs` matches nothing, and a level that *is* in `dfCovs` but not in the map
falls through to its raw code (`SEX=2`) in the finished plot. The template can
only list what is there, so both mistakes are visible before anything is
plotted.

Filled in, it gives the three maps below.

```{r plot_forest, fig.height=20, fig.width=16, out.width="100%"}
# 1. The maps, filled in from the template above
unit_map <- c(
  AGE = "years", WT = "kg", BMI = "kg/m2", BSA = "m2",
  CRCL = "mL/min", BILI = "mg/dL", AST = "IU", ALT = "IU",
  HT = "cm", LBM = "kg"
)

label_map <- c(
  "SEX=1" = "Male", "SEX=2" = "Female",
  "RACEL=1" = "Caucasian", "RACEL=2" = "Black", "RACEL=3" = "Other",
  "SMOK=0" = "Non-smoking", "SMOK=1" = "Smoking",
  "NCIL=0" = "NCI Grade 0", "NCIL=1" = "NCI Grade 1", "NCIL=2" = "NCI Grade 2",
  "GENO2=1" = "Poor metab.", "GENO2=2" = "Extensive metab."
)

group_map <- c(
  "AGE" = "Age", "ALT" = "ALT", "AST" = "AST", "BILI" = "Total Bilirubin",
  "BMI" = "BMI", "BSA" = "BSA", "CRCL" = "Creatinine CL",
  "ETHNIC" = "Ethnicity", "GENO2" = "Genotype", "HT" = "Height",
  "LBWT" = "Lean Body Wt", "NCIL" = "NCI", "RACEL" = "Race",
  "SEX" = "Sex", "SMOK" = "Smoking Status", "WT" = "Weight"
)

# 2. Generate the labels and groups from dfCovs, and attach them to the results
dfres_formatted <- generateCovNames(
  dfCovs,
  dfres = dfres,
  unit_map = unit_map,
  label_map = label_map,
  group_map = group_map
)

# 3. Render Plot
PMXForest::forestPlot(
  dfres_formatted,
  tabTextSize = 14,
  strip_right_size = 9
)
```
