---
title: "2. Walkthrough: Complete FREM Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Walkthrough: Complete FREM Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

This Walkthrough builds upon the fundamentals introduced in the Quickstart vignette. It is designed for pharmacometricians constructing robust, reproducible FREM workflows. Here, we move beyond the "happy path" to address real-world execution: constructing a Full Fixed Effects Model (FFEM) for evaluation, visualizing EBEs and diagnostics, plotting Visual Predictive Checks (VPCs), formatting presentation-ready tables, and safely updating models.

### **1. Setup and Global Options**

We begin by loading the necessary packages and setting our global plotting standard to `theme_bw()`. 

```{r setup}
library(PMXFrem)
library(ggplot2)
library(tidyverse)
library(kableExtra)

# Enforce standard ggplot theme
theme_set(theme_bw())

# Define paths to example data and converged models
modDevDir <- system.file("extdata", "SimNeb", package = "PMXFrem")
base_data <- system.file("extdata", "SimNeb/DAT-2-MI-PMX-2-onlyTYPE2-new.csv", package = "PMXFrem")

# Create a temporary directory for all generated outputs
td <- tempdir()
```

---

### **2. The FFEM Conversion**

Because a FREM model estimates covariates as dependent variables, it is mathematically incapable of directly producing standard diagnostics or VPCs. To evaluate the model, we must convert it into a deterministic Full Fixed Effects Model (FFEM). 

The `createFFEMmodel()` function locks the covariates into the dataset as individual parameter coefficients. We execute this to create the `.mod` and `.csv` files, saving them directly to our temp directory.

```{r create_ffem}
ffemMod <- createFFEMmodel(
  runno            = 31,
  baserunno        = 30,
  modDevDir        = modDevDir,
  numNonFREMThetas = 7,
  numSkipOm        = 2,
  parNames         = c("CL", "V", "MAT"),
  dataFile         = base_data,
  newDataFile      = file.path(td, "ffem_data.csv"),
  ffemModName      = file.path(td, "run31max0.mod") 
)
```

---

### **3. Model Evaluation: GOF Diagnostics**

Standard goodness-of-fit plots must be generated from the FFEM output. Here we read the standard table file generated by the FFEM to evaluate the model's structural fit.

```{r gof_plots, fig.width=10, fig.height=5}
# Read the pre-computed FFEM table
ffem_tab <- read.table(file.path(modDevDir, "xptabmax031"), skip = 1, header = TRUE) %>%
  filter(WRES != 0, BLQ != 1) # Keep only valid observation rows

# 1. PRED vs DV
p_pred <- ggplot(ffem_tab, aes(x = PRED, y = DV)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(title = "DV vs. Population Prediction (FFEM)", x = "PRED", y = "DV")

# 2. CWRES vs TIME
p_cwres <- ggplot(ffem_tab, aes(x = TIME, y = CWRES)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(title = "Conditional WRES vs. Time", x = "Time (h)", y = "CWRES")

# Combine plots
gridExtra::grid.arrange(p_pred, p_cwres, ncol = 2)
```

---

### **4. Model Evaluation: ETAs vs EBEs**

A critical conceptual difference in FREM is understanding the individual `ETA`s.

* In the **FREM** model, the raw `ETA` values contain both the unexplained variability *and* the covariate variability. They will show strong trends when plotted against covariates.
* In the **FFEM** model, the `EBE`s (Empirical Bayes Estimates) represent the unexplained variability *after* the covariates have been applied. 

Let's read the `.phi` files from both models using the internal `getPhi` function and plot them against Weight to visualize this difference.

```{r eta_vs_ebe, fig.width=10, fig.height=5}
# Extract ETA3 (CL) from the FREM and FFEM phi files
frem_phi <- getPhi(file.path(modDevDir, "run31.phi"))
ffem_phi <- getPhi(file.path(modDevDir, "run31max0.phi"))

# Fetch WT from the base data
subj_data <- read.csv(base_data) %>% distinct(ID, .keep_all = TRUE)

# Merge
eta_df <- data.frame(
  ID = subj_data$ID,
  WT = subj_data$WT,
  FREM_ETA = frem_phi$ETA.3.,
  FFEM_EBE = ffem_phi$ETA.3.
) %>% filter(WT != -99) # Remove missing baseline covariate rows for clean plotting

# Plot FREM 
p_frem <- ggplot(eta_df, aes(x = WT, y = FREM_ETA)) +
  geom_point(color = "darkorange", alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, color = "black") +
  labs(title = "FREM ETA vs WT", x = "Weight (kg)", y = "Base ETA (CL)")

# Plot FFEM 
p_ffem <- ggplot(eta_df, aes(x = WT, y = FFEM_EBE)) +
  geom_point(color = "steelblue", alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, color = "black") +
  labs(title = "FFEM EBE vs WT", x = "Weight (kg)", y = "EBE (CL)")

gridExtra::grid.arrange(p_frem, p_ffem, ncol = 2)
```
*Notice how the strong trend in the FREM ETA is completely resolved in the FFEM EBE, indicating that the FREM covariance matrix successfully captured the weight effect!*

---

### **5. Visual Predictive Checks (VPC)**

Predictive performance must also be simulated using the FFEM model so that the simulations condition on the actual observed covariates. We use the pre-computed `vpctab31` generated from the FFEM model.

```{r plot_vpc}

## VPC configuration
n_bins     <- 8
vpc_q      <- c(p10 = 0.10, p50 = 0.50, p90 = 0.90)
ci_q       <- c(lower = 0.05, median = 0.50, upper = 0.95)
vpc_colors <- c(p10 = "#005B65", p50 = "#E1A178", p90 = "#005B65")

calculate_bins <- . %>%
  group_by(time_bin) %>%
  summarise(
    x_time = median(TAD),
    p10 = quantile(DV, probs = vpc_q["p10"]),
    p50 = quantile(DV, probs = vpc_q["p50"]),
    p90 = quantile(DV, probs = vpc_q["p90"]),
    .groups = "drop" # Suppress grouping messages
  )

# Read the VPC data
vpctab <- read.table(file.path(modDevDir, "vpctab31"), skip = 1, header = TRUE)

## Process simulated data
vpc_sim_summary <- vpctab %>%
  filter(TAD != 0) %>%
  mutate(DV = exp(DV), time_bin = cut_number(TAD, n = n_bins)) %>%
  group_by(REPI) %>%
  nest() %>%
  mutate(binned = purrr::map(data, calculate_bins)) %>%
  unnest(binned) %>%
  pivot_longer(
    cols = starts_with("p"), 
    names_to = "percentile", 
    values_to = "value"
  ) %>%
  # Calculate confidence intervals across replicates
  group_by(time_bin, x_time, percentile) %>%
  summarise(
    lower = quantile(value, probs = ci_q["lower"]),
    upper = quantile(value, probs = ci_q["upper"]),
    .groups = "drop"
  )

# Process observed data
vpc_obs_summary <- ffem_tab %>%
  filter(TAD != 0) %>%
  mutate(DV = exp(DV), time_bin = cut_number(TAD, n = n_bins)) %>%
  calculate_bins() %>%
  pivot_longer(
    cols = starts_with("p"), 
    names_to = "percentile", 
    values_to = "value")

## Plotting
vpc1 <- ggplot() +
  # Confidence intervals from simulated data
  geom_ribbon(
    data = vpc_sim_summary,
    aes(x = x_time, ymin = lower, ymax = upper, fill = percentile),
    show.legend = FALSE, alpha = 0.3
  ) +
  # Lines for observed data percentiles
  geom_line(
    data = vpc_obs_summary,
    aes(x = x_time, y = value, color = percentile),
    linetype = "dashed", linewidth = 1, show.legend = FALSE
  ) +
  # Map colors and fills manually
  scale_color_manual(values = vpc_colors) +
  scale_fill_manual(values = vpc_colors) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Dependent Variable (DV)"
  ) 

vpc1
```

---

### **6. Parameter and Covariate Reporting**

`PMXFrem` includes functions to extract presentation-ready tables. By setting `uncertainty = "CI"` and `includeShrinkage = TRUE`, the function integrates the FFEM model's shrinkage into the main structural parameter table. 

We use `kableExtra` to format the output.

```{r parameter_tables}
parTableData <- fremParameterTable(
  runno            = 31,
  modDevDir        = modDevDir,
  thetaNum         = 2:5,
  omegaNum         = c(1,3,4,5),
  sigmaNum         = 1,
  thetaLabels      = c("CL (L/h)", "V (L)", "MAT (h)", "D1 (h)"),
  omegaLabels      = c("IIV on RUV", "IIV on CL", "IIV on V", "IIV on MAT"),
  parNames         = c("CL", "V", "MAT"), 
  sigmaLabels      = c("RUV"),
  includeRSE       = TRUE,
  uncertainty      = "CI",             # Generate 95% CIs
  includeShrinkage = TRUE,             # Include Eta Shrinkage
  ffemModName      = "run31max0",      # Used to calculate true EBE shrinkage
  numNonFREMThetas = 7,
  numSkipOm        = 2,
  availCov         = "all",
  quiet            = TRUE
)
```

#### **Structural Parameters**

The estimates for the fixed effects and `RUV` is taken directly from the FREM model output while the `IIV` terms (except the `IIV` on `RUV`) are adjusted for the impact of the covariates.

```{r print_par_table}
parTableData$parameterTable %>%
  mutate(Estimate = as.character(signif(Estimate, 3))) %>%
  select(-Type) %>% 
  kbl(caption = "Table 1. Structural Model Parameters, Uncertainty, and EBE Shrinkage") %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  kable_styling(position = "center")
```

#### **Covariate Coefficients**
```{r print_cov_table}
parTableData$coefficientTable_wide %>%
  kbl(caption = "Table 2. FREM Covariate Coefficients") %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  kable_styling(position = "center")
```

---

### **7. Model Revision (`updateFREMmodel`)**

If you need to iteratively add or remove covariates without restarting your entire base workflow, `updateFREMmodel()` handles the dataset restructuring and matrix rebuilding automatically.

The example below safely generates a new, modified dataset and control stream in our temporary directory:

```{r update_frem}
# To avoid modifying the read-only package directory, we copy the base model 
# and its .ext file to our temporary directory first.
file.copy(file.path(modDevDir, "run31.mod"), file.path(td, "run31.mod"))
file.copy(file.path(modDevDir, "run31.ext"), file.path(td, "run31.ext"))

newFrem <- updateFREMmodel(
  strFREMModel      = file.path(td, "run31.mod"),
  strFREMData       = file.path(modDevDir, "frem_dataset.dta"),
  strFFEMData       = base_data,
  cstrRemoveCov     = c("SEX"),       # Remove the SEX covariate from the model
  strID             = "ID",
  strNewFREMData    = file.path(td, "frem_dataset_noSEX.csv"),
  numNonFREMThetas  = 7,
  numSkipOm         = 2,
  bWriteData        = TRUE,
  bWriteMod         = TRUE,
  quiet             = TRUE,
  cstrKeepCols      = c("ID", "TIME", "AMT", "EVID", "RATE", "DV", "FOOD", "FREMTYPE")
)
```
