---
title: "4. Deep Dive: Explained Variability"
author: "PMXFrem Development Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Deep Dive: Explained Variability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8,
  warning = FALSE,
  message = FALSE
)
```

## Introduction
This vignette explains how to use the `PMXFrem` package to visualize the degree to which covariates manage to explain variability in the parameters of interest. 

The workflow is divided into five logical steps:

1. **Pre-requisite Setup:** Loading models and datasets.
2. **Covariate Scenarios:** Defining the specific covariate combinations to isolate.
3. **Parameter Function Definition:** Writing the structural equations to evaluate.
4. **Computing Explained Variability:** Deriving the variance metrics using specific integration methods.
5. **Visualization:** Rendering the explained variability bar plots.

## 1. Setting up the Pre-requisite

Load the package and identify which NONMEM FREM run is going to be used.

```{r setup}
library(PMXFrem) 
library(ggplot2)
library(dplyr)
library(data.table)

# Enforce a clean global theme
set.seed(865765)
theme_set(theme_bw(base_size = 14))

# Specify the name of the FREM run and collect file paths
runname <- "run31"
modDevDir <- system.file("extdata/SimNeb", package = "PMXFrem")
fremFiles <- getFileNames(modName = runname, modDevDir = modDevDir)

numNonFREMThetas <- 7
numSkipOm        <- 2

dfData <- fread(system.file("extdata/SimNeb/DAT-2-MI-PMX-2-onlyTYPE2-new.csv", package="PMXFrem"), data.table = FALSE) %>% 
  distinct(ID, .keep_all = TRUE)
```

## 2. Defining the Covariate Scenarios

We use the package's internal `setupDfCovsEV()` function to dynamically construct the covariate isolation matrix. 

**Important Conceptual Note:** There is a fundamental difference in how the `dfCovs` matrix is utilized here compared to Forest plots. In a Forest plot, `dfCovs` dictates the exact numerical covariate values to visualize (e.g., `WT = 115`). For explained variability, `dfCovs` acts purely as an indicator matrix **for the FREM covariates**. For these covariates, the actual numerical value assigned does not impact the calculation; setting it to anything other than the missing covariate indicator (typically `-99`) simply flags *which* covariates should be actively evaluated for that specific scenario row.

**However, this indicator system only applies to FREM covariates.** For FFEM covariates (covariates outside the FREM matrix, such as `FOOD`), the parameter function will still evaluate the *actual numerical value* provided in `dfCovs`. 

Because `FOOD` is a time-varying categorical FFEM covariate that our parameter function evaluates manually, we append it as an `additionalCov` and set its state to `1` (Fed) so our function logic executes correctly.

```{r cov_setup}
# Robustly generate the covariate matrix
dfCovs <- setupDfCovsEV(
  modFileName = fremFiles$mod, 
  additionalCovs = c("FOOD")
)

# Lock the time-varying FFEM covariate state
dfCovs$FOOD <- 1

# Define the exact names corresponding to the rows in dfCovs
cstrCovariates <- c("All", names(dfCovs))
```

## 3. Defining the Parameter Function

We must supply a function that mimics the structural relationships in the FFEM model. This function is conceptually identical to the one used for Forest plots, with one essential difference: it **must** include the `etas[]` argument to allow the package to integrate over the structural variance.

**Handling Missing Covariates:** You must build defensive fallback logic for missing covariates (e.g., `dfrow$FOOD != -99`) *only* for covariates outside the FREM part of the model (FFEM covariates). `PMXFrem` automatically handles missing data for the covariates included in the FREM covariance matrix. However, because the `dfCovs` matrix intentionally assigns `-99` to isolate covariate effects, your function will encounter missing data for FFEM covariates like `FOOD` and must know how to revert to the baseline fixed effect.

```{r define_function}
paramFunctionExpVar <- function(basethetas, covthetas, dfrow, etas, ...) {

  # Defensive fallback logic for isolating FFEM covariate effects
  FRELFOOD <- 1
  if (any(names(dfrow) == "FOOD") && dfrow$FOOD != -99 && dfrow$FOOD == 0) {
    FRELFOOD <- (1 + basethetas[7])
  }

  MATFOOD <- 1
  if (any(names(dfrow) == "FOOD") && dfrow$FOOD != -99 && dfrow$FOOD == 0) {
    MATFOOD <- (1 + basethetas[6])
  }

  TVFREL <- basethetas[1]
  TVCL   <- basethetas[2]
  TVV    <- basethetas[3]
  TVMAT  <- basethetas[4]
  TVD1   <- basethetas[5]

  MU_2 <- TVD1
  MU_3 <- log(TVFREL)
  MU_4 <- log(TVCL)
  MU_5 <- log(TVV)
  MU_6 <- log(TVMAT)

  D1FR  <- MU_2 + etas[2]
  FREL  <- FRELFOOD * exp(MU_3)
  CL    <- exp(MU_4 + covthetas[1] + etas[3])
  V     <- exp(MU_5 + covthetas[2] + etas[4])
  MAT   <- MATFOOD * exp(MU_6 + covthetas[3] + etas[5])

  D1   <- MAT * (1 - TVD1)
  KA   <- 1 / (MAT - D1)

  # Compute AUC after an 80 mg dose
  AUC <- 80 / (CL / FREL)

  return(list(CL, FREL, AUC, V, MAT))
}

functionListName <- c("CL", "Frel", "AUC", "V", "MAT")
```

## 4. Computing Explained Variability

The `getExplainedVar()` function computes the total and explained variability. The variability can be related to the total variability in the parameter ($TOTVAR$) or the total covariate variability in the parameter ($TOTCOVVAR$).

The `dfres` output contains three main metrics:

* $TOTVAR$: The total variance, calculated using the Delta method (Type 0), EBEs (Type 1), or $\eta$ samples (Type 2, 3).
* $TOTCOVVAR$: Calculated using all covariates with $\eta=0$, across all observed covariate values.
* $COVVAR$: Calculated with one (or a combination) of covariates, setting $\eta=0$.

### Calculation Methods (`type`)
The methodology is dictated by the `type` argument, and each has distinct mathematical properties:

* **Type 0 (Delta Method):** Based on omega calculations. It is fast and guarantees that the explained variability of all covariates will sum to exactly 100%.
* **Type 1 (EBEs):** Uses Empirical Bayes Estimates. It is subject to $\eta$ shrinkage but acknowledges individual information richness.
* **Type 2 ($\eta$ Samples):** Uses $\eta$ samples and integrates over observed FFEM covariates. Useful if the model contains non-FREM covariates.
* **Type 3 ($\eta$ Samples Optimized):** Similar to Type 2 but does not acknowledge FFEM covariates, making it significantly faster.

*Note: With Types 1, 2, and 3, it is theoretically possible for a single covariate to explain more than all covariates combined (i.e., > 100%) due to parameter correlations and structural non-linearities.*

```{r compute_var}
dfres <- getExplainedVar(
  type             = 2,
  data             = dfData,
  dfCovs           = dfCovs,
  numNonFREMThetas = numNonFREMThetas,
  numSkipOm        = numSkipOm,
  functionList     = list(paramFunctionExpVar),
  functionListName = functionListName,
  cstrCovariates   = cstrCovariates,
  modDevDir        = modDevDir,
  modName          = runname,
  ncores           = 1,
  numETASamples    = 100,
  quiet            = TRUE,
  seed             = 123
)
```

## 5. Visualization

We plot the results using `plotExplainedVar()`. To ensure reproducibility and prevent silent misalignment, we map our labels using a programmatic dictionary.

```{r plot_var}
# 1. Define robust label mappings
label_map <- c(
  "All" = "All Covariates", "AGE" = "Age", "ALT" = "ALT", "AST" = "AST", 
  "BILI" = "Bilirubin", "BMI" = "BMI", "BSA" = "BSA", "CRCL" = "Creatinine Clearance", 
  "ETHN" = "Ethnicity", "GENO2" = "Genotype", "HT" = "Height", "LBM" = "Lean Body Mass", 
  "NCIL" = "NCI Grade", "RACEL" = "Race", "SEX" = "Sex", "SMOK" = "Smoking Status", 
  "WT" = "Weight", "FOOD" = "Food Status"
)

# 2. Safely translate cstrCovariates into clean labels
mapped_labels <- unname(sapply(cstrCovariates, function(x) {
  ifelse(x %in% names(label_map), label_map[[x]], x)
}))

# 3. Plot Total Explained Variability (maxVar = 1)
# Visualizes the fraction: 100 * (COVVAR / TOTVAR)
plotExplainedVar(
  dfres, 
  parameters = "AUC", 
  maxVar = 1, 
  covariateLabels = mapped_labels
) + labs(title = "Explained Part of Total Variability (AUC)")

# 4. Plot Proportion of Explainable Variability (maxVar = 2)
# Visualizes the fraction: 100 * (COVVAR / TOTCOVVAR)
plotExplainedVar(
  dfres, 
  parameters = "AUC", 
  maxVar = 2, 
  covariateLabels = mapped_labels
) + labs(title = "Explained Part of Explainable Variability (AUC)")
```