---
title: "3. Deep Dive: Creating FREM Models"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Creating FREM Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The Full Random Effects Model (FREM) methodology requires a specific restructuring of both the NONMEM control stream and the underlying dataset. Continuous and categorical covariates must be converted into dummy "observations" so their variability and correlation can be estimated alongside the structural parameters via a unified covariance matrix.

`createFREMmodel()` does that transformation: it takes a base model and its data set and writes the FREM control stream and data set NONMEM needs.

# How createFREMmodel Works Under the Hood
Rather than adding every covariate at once, `createFREMmodel()` builds the model one covariate at a time. It first writes a minimal FREM model from the first covariate in your `covariates` argument, rewriting the `$INPUT`, `$DATA`, `$PK` and `$ERROR` blocks for that one covariate. It then works through the rest, calling `updateFREMmodel()` for each, so every step is the same operation the update functions already perform.

# 1. Setup and Paths

First, load the package and define the paths to your base model and original clinical dataset. 

`createFREMmodel()` writes into `modDevDir` by default, and stops rather than overwrite an existing file with the same `fremModName`. This vignette writes into a temporary directory instead.

```{r setup}
library(PMXFrem)

# Define paths to example data and converged models
modDevDir <- system.file("extdata", "SimNeb", package = "PMXFrem")
ffemDataFile <- system.file("extdata", "SimNeb/DAT-2-MI-PMX-2-onlyTYPE2-new.csv", package = "PMXFrem")

# Create a temporary directory for all generated outputs
td <- tempdir()
```

# 2. Designing the FREM Model

We need to decide which covariates to include and how they should be handled. In FREM, continuous covariates can be modeled as-is or log-transformed, while categorical covariates must be identified so they can be binarized into dummy variables.

We will also explicitly define `keep_cols` so that the columns the model needs carry over to the new FREM data set rows.

```{r define_covariates}
# Define the complete list of covariates to include
frem_covariates <- c("AGE", "WT", "SEX", "RACEL")

# Identify categorical and log-transformed covariates
categorical_covs <- c("RACEL", "SEX")
log_transform_covs <- c("WT")

# Define which structural columns must be kept in the new FREM dataset rows
keep_cols <- c("ID", "TIME", "AMT", "EVID", "RATE", "FOOD", "DV")
```

# 3. Generating the Model

We pass our specifications into `createFREMmodel()`. We use the `run30` base model, which has `7` structural THETAs (`numNonFREMThetas`) and `2` specific OMEGAs that we want to skip correlating with the covariates (`numSkipOm`). `numNonFREMThetas` can be left `NULL` -- it is then read from the base model's `.ext` -- but `numSkipOm` cannot be deduced from a base model and must be given.

## Dichotomous covariates

`bRecodeDichotomous = TRUE` recodes two-level categorical covariates from
`1`/`2` - how many data sets code sex - to `0`/`1`, which makes `0` the
reference level and the covariate effect the difference from it. Without it,
FREM refuses a `1`/`2` coding: see **Changes to existing behaviour** in the
2.0.0 release notes.



```{r execute_create, warning=FALSE, message=FALSE}
# Execute the creation of the FREM model and dataset
generated_files <- createFREMmodel(
  modName = "run30",
  modDevDir = modDevDir,
  ffemDataFile = ffemDataFile,
  outputDir = td,
  fremModName = "run30_FREM",

  # Covariate Instructions
  covariates = frem_covariates,
  catCovs = categorical_covs,
  logtCovs = log_transform_covs,

  # Structural columns
  numNonFREMThetas = 7,
  numSkipOm = 2,

  # Dataset formatting
  cstrKeepCols = keep_cols,
  bRecodeDichotomous = TRUE,
  quiet = TRUE
)
```

# 4. Understanding the Output

`createFREMmodel()` has now generated the newly expanded `.mod` and `.csv` files. Let's inspect them directly from the disk.

## The Generated Dataset
The original dataset has been vertically expanded. For every subject, new rows (`EVID = 2`) have been appended containing the standardized covariate values in the `DV` column, mapped via the new `FREMTYPE` compartment column. Because we used `bRecodeDichotomous = TRUE`, `SEX` appears as `0` or `1`.

```{r show_data}
# Read the generated dataset
frem_data <- read.csv(generated_files$data)

# Display the first 15 rows to show the structural PK data followed by the new FREM dummy rows
head(frem_data, n = 15)
```

## The Final Model Control Stream
The control stream is substantially different. The `$PK` block now handles the structural `ETA`s alongside the covariate means, and the `$ERROR` block routes the dummy observations (via `FREMTYPE`) to their respective equations. Finally, a large `$OMEGA BLOCK` is appended to estimate the covariances.

```{r show_model, comment=''}
# Read the generated NONMEM control stream and print it in its entirety
frem_mod <- readLines(generated_files$model)
cat(frem_mod, sep = "\n")
```

`run30_FREM.mod` is ready to run in NONMEM.
