---
title: "3. Deep Dive: Subject-Specific Variances for Missing Covariates"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Subject-Specific Variances for Missing Covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The FFEM conversion normally derives one set of variance parameters and gives it
to every subject. That assumes everyone carries the same covariate information.
Where covariates are missing, and missing in different combinations from subject
to subject, the assumption does not hold: a subject whose weight was never
measured should keep the variability that weight would have explained, while a
subject with a full covariate record should not.

`omegaToData = TRUE` relaxes it. Each subject gets the variance-covariance
matrix implied by their own missing-covariate pattern, carried in the data set
rather than in `$OMEGA`.

# 1. How it works

The matrix a subject is given is the FREM covariance conditional on the
covariates that subject actually has. Since it differs between subjects it
cannot live in `$OMEGA`, which is estimated once for the run. Instead the
elements travel in the data set, and the model rebuilds the correlated random
effects from them through a Cholesky factorization of the individual matrix
\(V\):

$$V = L L^{T}$$

\(L\) is lower triangular, so independent standard normal `ETA`s become
correlated `MYETA`s by

$$L_{ii} = \sqrt{V_{ii} - \sum_{k<i} L_{ik}^{2}}, \qquad
  L_{ij} = \frac{V_{ij} - \sum_{k<j} L_{ik} L_{jk}}{L_{jj}}$$

The indices are the model's own eta numbers, so in a model with skipped omegas
they start above them: `run31` skips two, and its FREM parameter etas are 3, 4
and 5.

# 2. The data set

`createFFEMdata(omegaToData = TRUE)` appends one column per element of the lower
triangle, named for the etas it belongs to.

```{r ffem_data}
library(PMXFrem)

modDevDir <- system.file("extdata/SimNeb", package = "PMXFrem")
td <- tempdir()

ffem <- createFFEMdata(
  modName     = "run31",
  modDevDir   = modDevDir,
  parNames    = c("CL", "V", "MAT"),
  dataFile    = file.path(modDevDir, "DAT-2-MI-PMX-2-onlyTYPE2-new.csv"),
  newDataFile = file.path(td, "ffem_omegaToData.csv"),
  omegaToData = TRUE,
  quiet       = TRUE
)

vCols <- grep("^V[0-9]+$", names(ffem$newData), value = TRUE)
vCols
```

`numNonFREMThetas` and `numSkipOm` are not passed: the model is located through
`modName` / `modDevDir`, so they come from the model and its `.ext` via
`fremModelInfo()`. `V33` is the variance of eta 3, `V43` its covariance with
eta 4, and so on.

The values differ from subject to subject:

```{r v_per_subject}
firstRow <- ffem$newData[!duplicated(ffem$newData$ID), ]
head(firstRow[, c("ID", vCols)], 4)
```

They differ because the covariate records differ. Subject 19 is missing one
covariate more than subject 1, and keeps a larger unexplained variance on
clearance:

```{r why_they_differ}
covs <- getCovNames(file.path(modDevDir, "run31.mod"))$orgCovNames
covs <- intersect(covs, names(firstRow))

nMissing <- function(id) sum(firstRow[firstRow$ID == id, covs] == -99)

data.frame(
  ID              = c(1, 19),
  covsMissing     = c(nMissing(1), nMissing(19)),
  varianceOnCL    = firstRow$V33[firstRow$ID %in% c(1, 19)]
)
```

# 3. The model code

`createFFEMmodel(omegaToData = TRUE)` fixes the FREM part of `$OMEGA` to an
identity matrix and writes the factorization into `$PK`, reading the `V`-columns
from the data set.

```{r ffem_model}
ffemMod <- createFFEMmodel(
  runno       = 31,
  baserunno   = 30,
  modDevDir   = modDevDir,
  parNames    = c("CL", "V", "MAT"),
  dataFile    = file.path(modDevDir, "DAT-2-MI-PMX-2-onlyTYPE2-new.csv"),
  newDataFile = file.path(td, "ffem_omegaToData_model.csv"),
  ffemModName = file.path(td, "run31_omegaToData.mod"),
  omegaToData = TRUE,
  quiet       = TRUE
)
```

The generated `$PK` block:

```{r cholesky_block, comment = ""}
i <- grep("Cholesky factors", ffemMod)
cat(ffemMod[i:(i + 21)], sep = "\n")
```

And the `$OMEGA` records the factorization replaces. The two skipped omegas are
untouched; the FREM block is an identity matrix, fixed:

```{r omega_block, comment = ""}
j <- grep("^\\$OMEGA", ffemMod)[1]
cat(ffemMod[j:(j + 6)], sep = "\n")
```

# 4. When to use it

`omegaToData = TRUE` is worth the extra columns when the covariate record is
incomplete and the gaps are uneven - a subject missing four covariates and one
missing none should not share a variance. With a complete covariate record every
subject gets the same matrix, and the default `omegaToData = FALSE` says the
same thing with less machinery.
