---
title: "3. Deep Dive: Reconciling FREM and SCM Forest Plots"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Reconciling FREM and SCM Forest Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

# Introduction

A Forest plot of the same covariate, drawn from an SCM model and from a FREM
model of the same data, can point in opposite directions without either being
wrong. They answer different questions, and the difference is not in the
estimation but in two choices made when the plot is drawn: what the predictions
are conditioned on, and what they are compared against.

This vignette shows which question each plot answers, and how to make one FREM
model produce either.

# 1. Conditional and unconditional covariate effects

**A conditional covariate effect** is the isolated effect of one covariate with
every other covariate held fixed. It supports a mechanistic reading - what this
covariate does, once the others are accounted for - and it requires a
multivariable model.

**An unconditional covariate effect** is the total effect of a covariate,
whatever else it travels with in the population. It supports label statements
and dosing recommendations - what to expect of a subject known only by this one
covariate - and it is what a univariable model estimates.

An SCM produces one multivariable model, so its Forest plot is conditional. FREM
estimates the covariates jointly and can be reduced to any subset of them, so a
single FREM model can answer either question. Which one it answers depends on
how the plot is set up.

# 2. What the two defaults do

|  | SCM | FREM (default) |
|---|---|---|
| Predictions | conditional on all other covariates | unconditional |
| Reference | a typical subject as the model is parameterised - centred on the median, or the most common category | a typical subject as the *data* defines it: the mean predicted parameter value across the covariates |

Neither default is the "right" one. They differ because the two methods were
built for different readings, which is why the same covariate can sit on
opposite sides of 1 in the two plots.

# 3. The two controls

Everything below is done with two inputs, and both are built by PMXForest rather
than assembled by hand:

* **`dfCovs`**, from `setupDfCovs()` - one row per Forest plot row. The
  covariates active in a row are the ones the prediction is conditioned on. With
  `useMissVal = TRUE` (the default) every other covariate is left at `missVal`
  and only one is active per row; with `useMissVal = FALSE` the inactive ones
  are set to their reference value instead, which conditions every row on them.
* **`dfRefRow`**, from `setupDfRefRow()` - the covariate values the predictions
  are compared against. `singleRef = TRUE` returns one row, and that same
  comparator is used by every row of the plot. `singleRef = FALSE` returns a
  data frame with as many rows as `dfCovs`, so each row of the plot gets its own
  comparator. Passing no `dfRefRow` at all leaves the FREM default in place.

Both take `contRef` and `catRef` to say what "reference" means for each
covariate - a number, `"mean"`, `"median"`, `"mode"`, `"lowest"`, or `"model"`
to read it from the control stream. The defaults are the median of a continuous
covariate and the most common level of a categorical one.

The other choice is the scale: `plotRelative = TRUE` plots each row as a ratio
to its reference, `FALSE` on the parameter's own scale.

# 4. Setup

```{r setup}
library(PMXFrem)
library(PMXForest)
library(ggplot2)
library(data.table)

set.seed(865765)
theme_set(theme_bw(base_size = 14))

modDevDir <- system.file("extdata/SimNeb", package = "PMXFrem")
fremFiles <- getFileNames(modName = "run31", modDevDir = modDevDir)

data <- fread(
  file.path(modDevDir, "DAT-2-MI-PMX-2-onlyTYPE2-new.csv"),
  data.table = FALSE
)

covariates <- c("AGE", "GENO2", "RACEL", "WT")
```

Four covariates are enough to make the point, and keep the plots readable:

```{r dfcovs}
dfCovs <- setupDfCovs(
  data       = data,
  covariates = covariates,
  probs      = c(0.05, 0.9),
  idVar      = "ID",
  missVal    = -99
)

dfCovs
```

Each row sets one covariate and leaves the rest at `missVal`. What that means
depends on the model behind the parameter function. In a **FREM** model the
covariates left at `missVal` are integrated out, so the prediction is
unconditional. In an **SCM** or FFEM parameter function there is nothing to
integrate over: `missVal` triggers the model's own missing-covariate handling,
which puts the covariate at its reference - so the prediction is conditioned on
the reference values rather than unconditional. Same `dfCovs`, two different
readings, which is where the two plots start to diverge.

```{r samples}
dfExt <- subset(getExt(extFile = fremFiles$ext), ITERATION == "-1000000000")

dfParameters <- getSamples(fremFiles$cov, extFile = fremFiles$ext, n = 100)
dfParameters[1, ] <- dfExt[1, -1] # first sample = the final estimates

# CL is THETA(2) in run31, and the model's first FREM parameter
functionList <- list(
  function(basethetas, covthetas, ...) basethetas[2] * exp(covthetas[1])
)
```

`RACEL` is coded 1 Caucasian, 2 Black and 3 Asian, with level 1 as the
reference, so the one-hot columns `RACEL_2` and `RACEL_3` carry the other two.

For didactic purposes in this vignette, the default labels are kept in full. The
varying dimensions of the resulting plots are managed through the small wrapper
function below.

```{r plot_helper}
fp <- function(dfres, left = 30, relative = TRUE, ...) {
  forestPlot(dfres,
    parameters = "CL", plotRelative = relative,
    errbartabwidth = c(3, 1), ...
  ) +
    theme(plot.margin = margin(5.5, 5.5, 5.5, left))
}
```

# 5. The FREM default: unconditional

```{r default}
dfres_default <- getForestDFFREM(
  dfCovs           = dfCovs,
  runno            = 31,
  modDevDir        = modDevDir,
  dfParameters     = dfParameters,
  functionList     = functionList,
  functionListName = "CL",
  quiet            = TRUE
)

data.frame(
  row       = dfres_default$COVNAME,
  CL        = round(dfres_default$POINT, 3),
  reference = round(dfres_default$REFFUNC, 3),
  relative  = round(dfres_default$POINT / dfres_default$REFFUNC, 3)
)
```

The same reference value for every row, and each row varies one covariate only.

```{r default_plot, fig.width = 10, fig.height = 5, out.width = "100%"}
fp(dfres_default)
```

# 6. Moving the reference

The reference need not be the one FREM computes from the data.
`setupDfRefRow()` builds one from the covariates themselves - here the typical
subject an SCM model would be centred on, the median of each continuous
covariate and the most common level of each categorical one:

```{r ref_single}
dfRefRow <- setupDfRefRow(
  dfCovs     = dfCovs,
  data       = data,
  covariates = covariates,
  singleRef  = TRUE
)

dfRefRow

dfres_typRef <- getForestDFFREM(
  dfCovs           = dfCovs,
  runno            = 31,
  modDevDir        = modDevDir,
  dfParameters     = dfParameters,
  functionList     = functionList,
  functionListName = "CL",
  dfRefRow         = dfRefRow,
  quiet            = TRUE
)

c(default = dfres_default$REFFUNC[1], typical = dfres_typRef$REFFUNC[1])
```

Every point estimate is unchanged - only what they are divided by has moved, so
the whole plot shifts. The predictions are still unconditional.

```{r ref_plot, fig.width = 10, fig.height = 5, out.width = "100%"}
fp(dfres_typRef,
  referenceInfo = "Reference: the typical subject - median AGE and WT, most common GENO2 and RACEL"
)
```

# 7. Conditioning the predictions: mimicking SCM

To read the plot the way an SCM plot is read, the predictions have to be
conditioned as well. That is `useMissVal = FALSE`: the covariates a row is not
varying are set to their reference instead of to `missVal`.

```{r scm_like}
dfCovsCond <- setupDfCovs(
  data        = data,
  covariates  = covariates,
  probs       = c(0.05, 0.9),
  idVar       = "ID",
  missVal     = -99,
  useMissVal  = FALSE
)

dfCovsCond

dfres_scm <- getForestDFFREM(
  dfCovs           = dfCovsCond,
  runno            = 31,
  modDevDir        = modDevDir,
  dfParameters     = dfParameters,
  functionList     = functionList,
  functionListName = "CL",
  dfRefRow         = dfRefRow,
  quiet            = TRUE
)

data.frame(
  row          = dfres_scm$COVNAME,
  relative_SCM = round(dfres_scm$POINT / dfres_scm$REFFUNC, 3)
)
```

Every row now carries every covariate, and the reference is that same typical
subject - which is what an SCM Forest plot shows. This is also visible in the
y-axis labels, kept here at their default: each covariate value that is not
`missVal` is included in the label.

```{r scm_plot, fig.width = 10, fig.height = 5, out.width = "100%"}
# the conditioned labels are long, so this one needs a wider margin
fp(dfres_scm,
  left = 95,
  referenceInfo = "Reference: the typical subject, with every prediction conditioned on it"
)
```

# 8. A reference that varies by covariate group

The unconditional reading is sharper when each covariate is compared against its
own reference rather than against one subject. That is `singleRef = FALSE`,
which returns a reference row per row of `dfCovs`:

```{r var_ref}
dfRefRowVarying <- setupDfRefRow(
  dfCovs     = dfCovs,
  data       = data,
  covariates = covariates,
  singleRef  = FALSE
)

dfRefRowVarying

dfres_varRef <- getForestDFFREM(
  dfCovs           = dfCovs,
  runno            = 31,
  modDevDir        = modDevDir,
  dfParameters     = dfParameters,
  functionList     = functionList,
  functionListName = "CL",
  dfRefRow         = dfRefRowVarying,
  quiet            = TRUE
)

data.frame(
  row       = dfres_varRef$COVNAME,
  reference = round(dfres_varRef$REFFUNC, 3),
  relative  = round(dfres_varRef$POINT / dfres_varRef$REFFUNC, 3)
)
```

Each covariate group now sits at 1 for its own reference level:

```{r var_ref_plot, fig.width = 10, fig.height = 5, out.width = "100%"}
fp(dfres_varRef,
  referenceInfo = "Reference varies by covariate: the median or most common level of each"
)
```

On the relative scale the varying reference is invisible - every group is
divided by its own value, so every group has a line at 1. On the absolute scale
it can be seen: the reference moves between covariate groups.

```{r var_ref_abs, fig.width = 10, fig.height = 5, out.width = "100%"}
fp(dfres_varRef,
  relative = FALSE,
  referenceInfo = "Reference varies by covariate: the median or most common level of each"
)
```

# 9. Which plot answers which question

The two readings can disagree, and on this model they do. Take Asian subjects -
`RACEL` level 3:

```{r estimands}
# the dfCovs row for Asian subjects. Match on the row number rather than
# the label: conditioning rewrites the label to carry every covariate it fixes.
iRacel3 <- which(dfCovs$COVARIATEGROUPS == "RACEL" & dfCovs$RACEL_3 == 1)

pick <- function(d) {
  r <- d[d$COVNUM == iRacel3, ]
  round(r$POINT / r$REFFUNC, 3)
}

c(unconditional = pick(dfres_varRef), conditional = pick(dfres_scm))
```

* **Unconditional**, `r pick(dfres_varRef)` of the reference: a subject known
  only to be Asian can be expected to have a *lower* CL than a Caucasian one.
  This is the reading for a label statement or a dosing recommendation.
* **Conditional**, `r pick(dfres_scm)` of the reference: an Asian subject who is
  otherwise typical can be expected to have a slightly *higher* CL than an
  otherwise identical Caucasian subject. This is the mechanistic reading, and
  the one an SCM Forest plot gives.

Neither number is wrong. They answer different questions, and a plot that does
not say which one it answers invites the reader to take the wrong one.
