---
title: "1. Quickstart: 5-Minute FREM Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Quickstart: 5-Minute FREM Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

This vignette provides a copy-pasteable example of the core **PMXFrem** workflow. We will take a converged FREM model, extract its parameters, visualize the covariate effects (Forest Plot), and calculate the explained variability. 

## 1. Load Package and Setup Paths
We use the example files provided within the package's `extdata` directory.


```{r setup}
library(PMXFrem)
library(ggplot2)
theme_set(theme_bw())

# Define paths to example data and models
modDevDir <- system.file("extdata", "SimNeb", package = "PMXFrem")
mod_path  <- file.path(modDevDir, "run31.mod")
ffemData  <- system.file("extdata", "SimNeb/DAT-2-MI-PMX-2-onlyTYPE2-new.csv", package = "PMXFrem")

# Setup a temporary directory for outputs (CRO Standard Practice)
td <- tempdir()

# Dynamically extract actual covariate names from the model (e.g., AGE, SEX, etc.)
cov_names <- getCovNames(mod_path)

```

## 2. Generate the FFEM Model
The primary entry point is `createFFEMmodel()`. This function automates the creation of the individual covariate coefficients and the corresponding NONMEM control stream.

```{r generate_ffem}
# Parameters in the model associated with covariates
parNames <- c("CL", "V", "MAT")

ffemMod <- createFFEMmodel(
  runno            = 31,
  baserunno        = 30,
  modDevDir        = modDevDir,
  dataFile         = ffemData,
  parNames         = parNames,
  numNonFREMThetas = 7,
  numSkipOm        = 2,
  newDataFile      = file.path(td, "vpcData31.csv"),
  ffemModName      = file.path(td, "run31_ffem.mod"),
  quiet            = TRUE
)

# Preview the first 10 lines of the generated NONMEM code
head(ffemMod, 10)
```

## 3. Extract Parameter Estimates
Once the FREM model has converged, we can extract the base parameters, adjusted variances (Omega prim), and residual errors (Sigma) using `fremParameterTable()`.

```{r param_table}
table_data <- fremParameterTable(
  runno            = 31,
  modDevDir        = modDevDir,
  numNonFREMThetas = 7,
  numSkipOm        = 2,
  thetaNum         = 2:7,
  omegaNum         = c(1, 3, 4, 5),
  sigmaNum         = 1,
  quiet            = TRUE
)

# View the formatted estimates
print(table_data$parameterTable)
```

## 4. Covariate Effects (Forest Plot)
To visualize the covariate effects, we first set up the covariate dataframe using our extracted names. We then pull bootstrap samples for uncertainty and plot using the  `PMXForest` package.

```{r forest_plot,, fig.asp = 2}

# 1. Setup the covariate matrix for the forest plot
data <- data.table::fread(ffemData)
dfCovsForest <- PMXForest::createInputForestData(
    PMXForest::getCovStats(data,cov_names$orgCovNames,probs=c(0.05,0.9),missVal= -99),
    iMiss    = -99
  )

# 2. Define the parameter function (No ETAs needed for the forest plot)
funcList_forest <- list(
  function(basethetas, covthetas, dfrow, ...) { 
    basethetas[2] * exp(covthetas[1]) 
  }
)

# 3. Get bootstrap samples for uncertainty
extFile <- file.path(modDevDir, "run31.ext")
bsFile  <- file.path(modDevDir, "bs31.dir/raw_results_run31.csv")

set.seed(123)
samples <- PMXForest::getSamples(bsFile, extFile = extFile, n = 25)

# 4. Generate covariate labels
formatted_labels <- generateCovNames(dfCovsForest)

# 5. Generate Forest Plot Data
dfresFREM <- getForestDFFREM(
  dfCovs           = dfCovsForest,
  covNames         = cov_names$covNames,
  cdfCovsNames     = formatted_labels,
  functionList     = funcList_forest,
  functionListName = "CL",
  numSkipOm        = 2,
  numNonFREMThetas = 7,
  dfParameters     = samples,
  quiet            = TRUE
)

# 5. Render Forest Plot
PMXForest::forestPlot(dfresFREM, parameters = "CL")
```

## 5. Visualizing Explained Variability
Finally, we calculate how much variability in our parameters is explained by these covariates. We use the Delta Method (`type = 0`) for rapid calculation.

```{r explain_var}

# 1. Setup the covariate matrix with actual covariate names
dfCovsEV <- setupDfCovsEV(mod_path, fremCovs = cov_names$orgCovNames)

# 1. Define the parameter function including ETAs for variability
funcList_var <- list(
  function(basethetas, covthetas, dfrow, etas, ...) { 
    basethetas[2] * exp(covthetas[1] + etas[3]) 
  }
)

# 2. Calculate variability
dfres_var <- getExplainedVar(
  type             = 0, 
  dfCovs           = dfCovsEV,
  cstrCovariates   =c("ALL",cov_names$orgCovNames),
  numNonFREMThetas = 7,
  numSkipOm        = 2,
  functionList     = funcList_var,
  functionListName = "CL",
  modDevDir        = modDevDir,
  runno            = 31,
  quiet            = TRUE
)

# 3. Plot results
plotExplainedVar(dfres_var)
```
