---
title: "3. Deep Dive: Reconciling FREM and SCM Forest Plots"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Reconciling FREM and SCM Forest Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Introduction

A Forest plot of the same covariate, drawn from an SCM model and from a FREM
model of the same data, can point in opposite directions without either being
wrong. They answer different questions, and the difference is not in the
estimation but in two choices made when the plot is drawn: what the predictions
are conditioned on, and what they are compared against.

This vignette shows which question each plot answers, and how to make one FREM
model produce either.

# 1. Conditional and unconditional covariate effects

**A conditional covariate effect** is the isolated effect of one covariate with
every other covariate held fixed. It supports a mechanistic reading - what this
covariate does, once the others are accounted for - and it requires a
multivariable model.

**An unconditional covariate effect** is the total effect of a covariate,
whatever else it travels with in the population. It supports label statements
and dosing recommendations - what to expect of a subject known only by this one
covariate - and it is what a univariable model estimates.

An SCM produces one multivariable model, so its Forest plot is conditional. FREM
estimates the covariates jointly and can be reduced to any subset of them, so a
single FREM model can answer either question. Which one it answers depends on
how the plot is set up.

# 2. What the two defaults do

|  | SCM | FREM (default) |
|---|---|---|
| Predictions | conditional on all other covariates | unconditional |
| Reference | a typical subject as the model is parameterised - centred on the median, or the most common category | a typical subject as the *data* defines it: the mean predicted parameter value across the covariates |

Neither default is the "right" one. They differ because the two methods were
built for different readings, which is why the same covariate can sit on
opposite sides of 1 in the two plots.

# 3. The three controls

Everything below is done with three arguments:

* **`dfCovs`** - one row per Forest plot row. The covariates set to something
  other than `missVal` in a row are the ones the prediction is conditioned on.
  The default, one covariate per row, gives unconditional predictions.
* **`dfRefRow`** - the covariate values the predictions are compared against.
  One row applies to every Forest plot row; a data frame with as many rows as
  `dfCovs` gives each row its own reference. `NULL` (the default) uses the
  typical subject.
* **`plotRelative`** and **`noVar`** in `forestPlot()` - the scale, and whether
  the uncertainty in the reference is carried into the interval (`noVar = FALSE`
  includes it). The absolute scale without that uncertainty is not a meaningful
  combination, and `forestPlot()` refuses it.

# 4. Setup

```{r setup}
library(PMXFrem)
library(PMXForest)
library(ggplot2)
library(data.table)

set.seed(865765)
theme_set(theme_bw(base_size = 14))

modDevDir <- system.file("extdata/SimNeb", package = "PMXFrem")
fremFiles <- getFileNames(modName = "run31", modDevDir = modDevDir)

data <- fread(
  file.path(modDevDir, "DAT-2-MI-PMX-2-onlyTYPE2-new.csv"),
  data.table = FALSE
)
```

Four covariates are enough to make the point, and keep the plots readable:

```{r dfcovs}
dfCovs <- setupDfCovs(
  data       = data,
  covariates = c("AGE", "GENO2", "RACEL", "WT"),
  probs      = c(0.05, 0.9),
  idVar      = "ID",
  missVal    = -99
)

dfCovs
```

Each row sets one covariate and leaves the rest at `-99`. That is what makes the
default predictions unconditional.

```{r samples}
dfExt <- subset(getExt(extFile = fremFiles$ext), ITERATION == "-1000000000")

dfParameters <- getSamples(fremFiles$cov, extFile = fremFiles$ext, n = 100)
dfParameters[1, ] <- dfExt[1, -1] # first sample = the final estimates

# CL is THETA(2) in run31, and the model's first FREM parameter
functionList <- list(
  function(basethetas, covthetas, ...) basethetas[2] * exp(covthetas[1])
)
```

# 5. The FREM default: unconditional

```{r default}
dfres_default <- getForestDFFREM(
  dfCovs           = dfCovs,
  runno            = 31,
  modDevDir        = modDevDir,
  dfParameters     = dfParameters,
  functionList     = functionList,
  functionListName = "CL",
  quiet            = TRUE
)

data.frame(
  row       = dfres_default$COVNAME,
  CL        = round(dfres_default$POINT, 3),
  reference = round(dfres_default$REFFUNC, 3),
  relative  = round(dfres_default$POINT / dfres_default$REFFUNC, 3)
)
```

One reference value for every row, and each row varies one covariate only.

```{r default_plot, fig.width = 9, fig.height = 5, out.width = "100%"}
forestPlot(dfres_default, parameters = "CL", plotRelative = TRUE, noVar = TRUE)
```

# 6. Moving the reference

The reference need not be the typical subject. `dfRefRow` takes the covariate
values to compare against - here the `GENO2` level with the higher clearance,
with every other covariate left at its average:

```{r ref_em}
covOnly <- function(d) d[, setdiff(names(d), "COVARIATEGROUPS"), drop = FALSE]

refEM <- covOnly(dfCovs[dfCovs$GENO2 == 2, ][1, ])
refEM

dfres_refEM <- getForestDFFREM(
  dfCovs           = dfCovs,
  runno            = 31,
  modDevDir        = modDevDir,
  dfParameters     = dfParameters,
  functionList     = functionList,
  functionListName = "CL",
  dfRefRow         = refEM,
  quiet            = TRUE
)

c(default = dfres_default$REFFUNC[1], GENO2_2 = dfres_refEM$REFFUNC[1])
```

Every point estimate is unchanged - only what they are divided by has moved, so
the whole plot shifts. The predictions are still unconditional.

# 7. Conditioning the predictions: mimicking SCM

To read the plot the way an SCM plot is read, two things change together: every
row is conditioned on the other covariates, and the reference becomes that same
typical subject.

```{r scm_like}
typical <- dfCovs
typical$AGE[typical$AGE == -99] <- 44
typical$WT[typical$WT == -99] <- 86
typical$GENO2[typical$GENO2 == -99] <- 2
typical$RACEL_2[typical$RACEL_2 == -99] <- 0
typical$RACEL_3[typical$RACEL_3 == -99] <- 0

refTypical <- covOnly(typical[1, ])
refTypical$AGE <- 44
refTypical$WT <- 86
refTypical$GENO2 <- 2
refTypical$RACEL_2 <- 0
refTypical$RACEL_3 <- 0

dfres_scm <- getForestDFFREM(
  dfCovs           = typical,
  runno            = 31,
  modDevDir        = modDevDir,
  dfParameters     = dfParameters,
  functionList     = functionList,
  functionListName = "CL",
  dfRefRow         = refTypical,
  quiet            = TRUE
)

data.frame(
  row          = dfres_scm$COVNAME,
  relative_SCM = round(dfres_scm$POINT / dfres_scm$REFFUNC, 3)
)
```

```{r scm_plot, fig.width = 9, fig.height = 5, out.width = "100%"}
forestPlot(dfres_scm, parameters = "CL", plotRelative = TRUE, noVar = TRUE,
           referenceInfo = "Reference: a 44-year-old subject of 86 kg, GENO2 = 2, RACEL reference level")
```

# 8. A reference that varies by covariate group

The unconditional reading is sharper when each covariate is compared against its
own typical value rather than against one subject. `dfRefRow` may carry as many
rows as `dfCovs`:

```{r var_ref}
refVarying <- covOnly(dfCovs)
refVarying$AGE <- ifelse(dfCovs$COVARIATEGROUPS == "AGE", 44, -99)
refVarying$WT <- ifelse(dfCovs$COVARIATEGROUPS == "WT", 86, -99)
refVarying$GENO2 <- ifelse(dfCovs$COVARIATEGROUPS == "GENO2", 2, -99)
refVarying$RACEL_2 <- ifelse(dfCovs$COVARIATEGROUPS == "RACEL", 0, -99)
refVarying$RACEL_3 <- ifelse(dfCovs$COVARIATEGROUPS == "RACEL", 0, -99)

dfres_varRef <- getForestDFFREM(
  dfCovs           = dfCovs,
  runno            = 31,
  modDevDir        = modDevDir,
  dfParameters     = dfParameters,
  functionList     = functionList,
  functionListName = "CL",
  dfRefRow         = refVarying,
  quiet            = TRUE
)

data.frame(
  row       = dfres_varRef$COVNAME,
  reference = round(dfres_varRef$REFFUNC, 3),
  relative  = round(dfres_varRef$POINT / dfres_varRef$REFFUNC, 3)
)
```

Each covariate group now sits at 1 for its own reference level. On the relative
scale that is not visible in the plot, so it belongs in the caption:

```{r var_ref_plot, fig.width = 9, fig.height = 5, out.width = "100%"}
forestPlot(dfres_varRef, parameters = "CL", plotRelative = TRUE, noVar = TRUE,
           referenceInfo = "Reference varies by covariate: AGE 44 y, WT 86 kg, GENO2 = 2, RACEL reference level")
```

# 9. Which plot answers which question

The two readings can disagree, and on this model they do. Take the third `RACEL`
level:

```{r estimands}
# the dfCovs row for the third RACEL level. Match on the row number rather than
# the label: conditioning rewrites the label to carry every covariate it fixes.
iRacel3 <- which(dfCovs$COVARIATEGROUPS == "RACEL" & dfCovs$RACEL_3 == 1)

pick <- function(d) {
  r <- d[d$COVNUM == iRacel3, ]
  round(r$POINT / r$REFFUNC, 3)
}

c(unconditional = pick(dfres_varRef), conditional = pick(dfres_scm))
```

* **Unconditional**, `r pick(dfres_varRef)` of the reference: a subject known
  only to be of that level can be expected to have a *lower* CL than one of the
  reference level. This is the reading for a label statement or a dosing
  recommendation.
* **Conditional**, `r pick(dfres_scm)` of the reference: a 44-year-old subject
  of 86 kg with `GENO2 = 2` and that level can be expected to have a slightly
  *higher* CL than an otherwise identical subject of the reference level. This
  is the mechanistic reading, and the one an SCM Forest plot gives.

Neither number is wrong. They answer different questions, and a plot that does
not say which one it answers invites the reader to take the wrong one.

# 10. Where explained variability fits

A Forest plot says how far a covariate moves a parameter. It does not say how
much of the variability in that parameter the covariate accounts for - a
covariate can move the parameter a long way and still explain little of the
spread, if few subjects are at its extremes. `getExplainedVar()` answers that
question, and the two plots are usually read together. See the Explained
Variability deep dive.
