---
title: "Deep Dive: Subject-Specific Variances for Missing Covariates (`omegaToData`)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deep Dive: Subject-Specific Variances for Missing Covariates (`omegaToData`)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Introduction and Clinical Rationale

In standard `createFREMmodel` executions (`omegaToData = FALSE`), only one set of variance parameters is derived for all individuals, operating under the assumption of no missing covariate values. 

However, in clinical datasets with significant missing data and high between-subject variability in missingness patterns, this assumption may fail. Setting `omegaToData = TRUE` relaxes this assumption. It ensures that each individual receives their own set of derived omega parameters, specifically adjusted according to their individual missing covariate pattern.

To achieve this, the PMXFrem package extracts the individualized variance-covariance elements directly into the dataset and inserts Cholesky decomposition mathematics into the NONMEM `$PK` or `$PRED` blocks.

## 2. Mathematical Foundation

When `omegaToData = TRUE`, PMXFrem circumvents the standard block-diagonal `$OMEGA` structure. Instead, it uses the Cholesky factorization of the individualized variance-covariance matrix $V$:

$$V = L L^T$$

Where $L$ is a lower-triangular matrix. Independent standard normal random variables ($\eta$) are mapped to correlated variables (`MYETA`) via linear combinations of the $L$ factors. For a $2 \times 2$ matrix, the inserted equations evaluate to:

$$L_{11} = \sqrt{V_{11}}$$
$$L_{21} = \frac{V_{21}}{L_{11}}$$
$$L_{22} = \sqrt{V_{22} - L_{21}^2}$$

## 3. Dataset Augmentation (`V`-Columns)

When invoking `createFFEMdata()` with `omegaToData = TRUE`, the matrix elements are appended to the output dataset as continuous `V`-columns. 

```{r eval=FALSE}
library(PMXFrem)

ffem_data <- createFFEMdata(
  modName = "run31",
  modDevDir = system.file("extdata/SimNeb/", package = "PMXFrem"),
  parNames = c("CL", "V"),
  numNonFREMThetas = 7,
  numSkipOm = 0,
  dataFile = my_data,
  omegaToData = TRUE
)
```

### Data Snippet
The resulting dataset assigns the individualized variances ($V_{11}, V_{22}$) and covariances ($V_{21}$) to each record based on that specific subject's missing covariate pattern:

```text
  ID TIME   DV       V11       V21       V22
1  1    0 10.5 0.0651800 0.0196800 0.0520700
2  1   12  8.2 0.0651800 0.0196800 0.0520700
3  2    0 20.1 0.0712000 0.0154000 0.0601000
4  2   24 15.0 0.0712000 0.0154000 0.0601000
```
*(Notice how Subject 1 and Subject 2 have different `V` values due to differing covariate missingness).*

## 4. Model Code Modifications

In `createFFEMmodel()`, activating this feature drops the empirical covariance estimates from the NONMEM `$OMEGA` block and replaces them with a fixed identity matrix. PMXFrem then inserts the structural equations directly into the model code.

```{r eval=FALSE}
ffem_model <- createFFEMmodel(
  runno = 31,
  baserunno = 30,
  numNonFREMThetas = 7,
  numSkipOm = 0,
  parNames = c("CL", "V"),
  dataFile = "raw_data.csv",
  omegaToData = TRUE
)
```

### Generated NONMEM Code Snippet
The generated model explicitly computes the $L$ factors using the data-supplied $V$ parameters and generates `MYETA` assignments representing the newly correlated random effects.

```text
$PK
; ----------------------------------------------------------------
; 1. Cholesky factors (L) for an 2x2 matrix
; ----------------------------------------------------------------

; Column 1
L11 = SQRT(MAX(0.000001, V11))
L21 = V21 / L11

; Column 2
L22 = SQRT(MAX(0.000001, V22 - L21**2))

; ----------------------------------------------------------------
; 2. Multiply standard ETA with L for correlated MYETA
; ----------------------------------------------------------------
MYETA1 = L11 * ETA(1)
MYETA2 = L21 * ETA(1) + L22 * ETA(2)

; Structural parameter assignment using MYETA and individualized covariate effects
CL = THETA(1) * EXP(MYETA1 + CLFREMCOV)
V  = THETA(2) * EXP(MYETA2 + VFREMCOV)

$OMEGA BLOCK(2) FIX
1.0
0.0 1.0
```