---
title: "3. Deep Dive: Secondary Parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Secondary Parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.callout-note {
  background-color: #e7f1ff;
  border-left: 4px solid #0dcaf0;
  padding: 12px;
  margin: 18px 0;
}
.callout-warning {
  background-color: #fff3cd;
  border-left: 4px solid #ffc107;
  padding: 12px;
  margin: 18px 0;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

```{r setup}
suppressPackageStartupMessages({
  library(PMXFrem)
  library(PMXForest)
  library(ggplot2)
  library(data.table)
})
theme_set(theme_bw(base_size = 14))
set.seed(865765)
```

# Introduction

A FREM Forest plot or explained-variability plot usually shows model parameters
-- `CL`, `V`, an absorption term. Often the quantity of clinical interest is a
**secondary** parameter derived from those: an exposure `AUC` over a dosing
interval, a terminal half-life, a steady-state `Cmax`.

`createFREMParamFunction()` transliterates the FREM model's `$PK` into an R
parameter function. Its `secondary` argument -- identical to
[PMXForest::createParamFunction()]'s -- attaches the derived quantities in the
same call, so the whole function (structural part and secondary part) is one
object that `getForestDFFREM()` and `getExplainedVar()` consume directly.

Each `secondary` entry names a variable to add to the return list; its value is
**one of**

* a line of R code -- `secondary = list(AUC = "dose / CL")`;
* the path to an `.R` file of arbitrary code, including an `mrgsolve`
  simulation -- `secondary = list(CMAX = "cmax.R")`; or
* a **list** with that `source` plus any constants the code needs --
  `secondary = list(CMAX = list(source = "cmax.R", dose = 100, tau = 12))`.
  Each constant is bound (`dose <- 100`, `tau <- 12`) immediately before the
  code, inside the same `local()` block. A bare string is just `source` with no
  constants.

The generated FREM function has the signature
`function(basethetas, covthetas, dfrow, etas = rep(0, ...), ...)`. A secondary
block sees `basethetas`, `covthetas`, `dfrow` (also aliased as `df`, so a
snippet written for the SCM version ports over), `etas`, `...`, any call-site
constants, and every structural parameter by name. **`etas = 0` gives the
forest-plot (typical) value; non-zero `etas` gives the value used by an
explained-variability plot** -- the same function serves both.

The two parts below are independent:

* **Part 1** derives `AUC` and a half-life in closed form. No compiled
  dependencies -- the code and output on this page are from a live run.
* **Part 2** computes a steady-state `Cmax` with an `mrgsolve` simulation. It is
  shown but **not executed**; the outputs are illustrative.

---

# Part 1 -- closed-form secondary parameters

## 1.1 Locate the FREM model

```{r locate}
runname   <- "run31"
modDevDir <- system.file("extdata/SimNeb", package = "PMXFrem")
fremFiles <- getFileNames(modName = runname, modDevDir = modDevDir)
```

## 1.2 Generate the parameter function with secondaries

`AUC` is a config list -- `source` is a one-liner and `dose` (100 mg) is
supplied at the call site. For a linear model the single-dose `AUC` is
`dose / CL`. `THALF` is a bare snippet that uses `CL` and `V`.

```{r genfun}
out <- createFREMParamFunction(
  fremModel  = fremFiles$mod,
  parameters = c("CL", "V", "MAT"),
  secondary  = list(
    AUC   = list(source = "dose / CL", dose = 100),
    THALF = "log(2) * V / CL"
  ),
  quiet = TRUE
)

out$functionListName   # CL, V, MAT, then the secondaries
out$primaryNames
out$secondaryNames
```

`numSkipOm` and `numNonFREMThetas` were derived from the model and its `.ext`
(via `fremModelInfo()`); nothing had to be passed by hand.

The tail of the generated source -- the constant binding, then each entry in
its own `local()` block, then the extended return list:

```{r gentail, echo = FALSE, comment = ""}
tail(out$code, 16)
```

## 1.3 Sanity-check at the final estimates

```{r sanity}
fun    <- eval(parse(text = out$code))
dfExt  <- subset(getExt(extFile = fremFiles$ext), ITERATION == "-1000000000")
thetas <- as.numeric(dfExt[1, grep("^THETA", names(dfExt))])
base   <- thetas[seq_len(out$noBaseThetas)]

fun(basethetas = base, covthetas = rep(0, out$numParCov), dfrow = data.frame(),
    etas = rep(0, out$numSkipOm + out$numParCov))
```

## 1.4 The Forest plot

`createFREMParamFunction()` returns one function that yields a named list;
`getForestDFFREM()` takes it as a one-element `functionList` with
`functionListName = out$functionListName`, and the covariate scenarios are
built the usual way. The model is located by `runno` / `modDevDir`, so
`numNonFREMThetas`, `numSkipOm` and `covNames` are all derived.

```{r forest}
covNames <- getCovNames(modFile = fremFiles$mod)

data   <- fread(
  system.file("extdata/SimNeb/DAT-2-MI-PMX-2-onlyTYPE2-new.csv", package = "PMXFrem")
)
dfCovs <- createInputForestData(
  getCovStats(data, covNames$orgCovNames, probs = c(0.05, 0.95), missVal = -99),
  iMiss = -99
)

dfParameters <- getSamples(fremFiles$cov, extFile = fremFiles$ext, n = 50)

dfres <- getForestDFFREM(
  dfCovs           = dfCovs,
  runno            = 31,
  modDevDir        = modDevDir,
  dfParameters     = dfParameters,
  functionList     = list(fun),
  functionListName = out$functionListName,
  quiet            = TRUE
)

unique(dfres$PARAMETER)
```

```{r forestplot, fig.width = 8, fig.height = 5, out.width = "100%"}
forestPlot(
  dfres[dfres$PARAMETER == "AUC", ],
  plotRelative = TRUE,
  keepYlabs    = TRUE,
  rightStrip   = FALSE
) +
  ggtitle("AUC vs. reference")
```

## 1.5 The same function in an explained-variability plot

`getExplainedVar()` calls the parameter function with **non-zero `etas`** to
propagate the FREM covariance. Because the secondary names are in
`functionListName`, `AUC` and `THALF` get an explained-variability row for
free -- no second function to write, no risk of the two drifting apart.

```{r ev, eval = FALSE}
dfCovsEV <- setupDfCovsEV(fremFiles$mod)

ev <- getExplainedVar(
  type             = 1,
  data             = data,
  dfCovs           = dfCovsEV,
  functionList     = list(fun),
  functionListName = out$functionListName,
  runno            = 31,
  modDevDir        = modDevDir,
  quiet            = TRUE
)
# ev now carries TOTVAR / covariate contributions for CL, V, MAT, AUC and THALF
```

---

# Part 2 -- a simulation-based secondary with mrgsolve

::: {.callout-warning}
**The code in this part is not executed when the vignette is built.** The
outputs shown are illustrative, not from a live run. The code *is* runnable, but
you must have a working `mrgsolve` installation first -- which needs a C++
toolchain (Rtools on Windows, the Xcode command-line tools on macOS,
`r-base-dev` / `build-essential` on Linux), because `mrgsolve` compiles each
model on first use.
:::

::: {.callout-note}
**mrgsolve version.** The example file uses only long-stable core API --
`mcode_cache()`, `param()`, `ev()`, `mrgsim_df()` -- and the basic
`$PARAM` / `$CMT` / `$ODE` / `$TABLE` / `$CAPTURE` blocks, so it needs **no
version-specific features**. Checked on **mrgsolve 1.0.9**; any
**mrgsolve >= 1.0.0** should be enough. The current CRAN release is **2.0.1**
and these calls are unchanged across the 1.x -> 2.x boundary, so the latest
version is *not* required.
:::

`PMXForest` ships a worked example at
`system.file("secondary-cmax-mrgsolve.R", package = "PMXForest")` -- a
one-compartment oral model that simulates a course of doses and returns the
steady-state `Cmax`. It is written against `df` (not `dfrow`), which is why
`createFREMParamFunction()` emits `df <- dfrow` at the top of the generated
function: the same file works from both `createParamFunction()` and
`createFREMParamFunction()`.

Pass it as a config-list `source`, with the regimen constants:

```r
cmaxFile <- system.file("secondary-cmax-mrgsolve.R", package = "PMXForest")

out <- createFREMParamFunction(
  fremModel  = fremFiles$mod,
  parameters = c("CL", "V", "MAT"),
  secondary  = list(
    CMAX = list(source = cmaxFile, dose = 80, tau = 24, n = 7)
  ),
  quiet = TRUE
)

out$functionListName
#> [1] "CL"   "V"    "MAT"  "CMAX"
```

The tail of the generated source carries the constants, then `df <- dfrow`, then
the file's text spliced in **verbatim** (re-indenting it would corrupt the
multi-line `mrgsolve` model string):

```r
  df <- dfrow   # alias for secondary code
  ...
  ## ---- secondary parameters ----
  ## CMAX  (inlined from secondary-cmax-mrgsolve.R)
  CMAX <- local({
    dose <- 80
    tau <- 24
    n <- 7
# ... file header ...
.mod <- mrgsolve::mcode_cache("pmxforest_secondary_1cmt_oral", .code)
# ...
.lastInterval <- subset(.sim, time >= (n - 1) * tau)
max(.lastInterval$CP)
  })

  list(CL = CL, V = V, MAT = MAT, CMAX = CMAX)
```

Evaluated at the final estimates for a reference subject:

```r
fun <- eval(parse(text = out$code))
fun(basethetas = base, covthetas = rep(0, out$numParCov), dfrow = data.frame(),
    etas = rep(0, out$numSkipOm + out$numParCov))
#> $CL
#> [1] ...
#> $CMAX
#> [1] 0.37          # ng/mL, illustrative
```

Then the Forest plot, exactly as in Part 1 -- `mrgsolve` must be loaded on the
worker processes, hence `cstrPackages`:

```r
dfres <- getForestDFFREM(
  dfCovs           = dfCovs,
  runno            = 31,
  modDevDir        = modDevDir,
  dfParameters     = dfParameters,
  functionList     = list(fun),
  functionListName = out$functionListName,
  cstrPackages     = c("PMXFrem", "mrgsolve"),
  quiet            = TRUE
)

forestPlot(dfres[dfres$PARAMETER == "CMAX", ], plotRelative = TRUE,
           keepYlabs = TRUE, rightStrip = FALSE)
```

`mrgsolve::mcode_cache()` keeps the compiled model on disk between calls; the
parameter function runs once per covariate row per sample, so recompiling every
call would dominate the run.

---

# Notes

* **Verification.** `verifyFREMParamFunction()` checks the structural part of the
  generated function against [PMXForest::createParamFunction()] on the same FREM
  model, plus the `covthetas` / `etas` scaling. It checks only the `$PK`
  parameters -- a secondary such as `AUC` has no structural counterpart -- so a
  function with secondaries still verifies cleanly.
* **Snippet vs. file vs. config list.** A string is a file when it exists on
  disk; a string that looks like a path (`.R` / `.r`) but does not exist is an
  error, not a silently mistyped snippet. A list entry needs a single-string
  `source`; every other element must be a named, atomic constant.
* **Constant names.** A call-site constant shadows a `$PK` variable of the same
  name inside that entry's `local()` block. Reusable files use dotted names
  (`.dose`) or an `exists(name, inherits = FALSE)` guard to stay clear of the
  structural namespace.
