---
title: "Revise an existing FREM model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manipulate an existing FREM model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,echo=FALSE}
suppressPackageStartupMessages(library(PMXFrem))
```

## Introduction

The PsN `frem` command generates a NONMEM model file that includes the FREM implementation of the covariates specified as input to `frem`. However, there may be instances when an existing FREM model has to be modified. Examples include the addition/removal of covariates from the model and the addition or removal of subjects in the analysis data set.

A particular use case may be when there is a lot of data and covariates and the FREM model from the `frem` command has convergence issues, or if the analyst wants to have more control of the addition of covariates and data sources to the model. 

Another model development challenge may be to update the initial estimates in the FREM model. The PsN command `update_inits` can usually be used for this but sometimes, with many covariates, the updated inits may not provide a valid OMEGA matrix for the FREM model.

A straight forward way to remove or add data or covariates to an existing FREM model is to re-run the PsN `frem` command with the new set of covariate. However, the `frem` command can sometime take a long time and the reason to generate a smaller FREM model in the first hand may be to add data and covariate in a granular fashion without involving the `frem` command.

The `PMXFrem` function `updateFREM()` addresses these issues by being able to add or remove covariates from an existing FREM model and/or to update the initial estimates of the parameters.


## Removing a covariate from a FREM model

```{r}
updateFREM(
  strFREMModel      = system.file("extdata/SimNeb/run31.mod", package = "PMXFrem"),
  strFREMData       = system.file("extdata/SimNeb/frem_dataset.dta", package = "PMXFrem"),
  strFFEMData       = system.file("extdata/SimNeb/DAT-2-MI-PMX-2-onlyTYPE2-new.csv", package = "PMXFrem"),
  cstrRemoveCov     = c("SEX"),
  cstrCatCovsToAdd  = NULL,
  cstrContCovsToAdd = NULL,
  strID             = "ID",
  strNewFREMData    = "frem_dataset_noSEX.csv",
  numNonFREMThetas  = 7,
  numSkipOm         = 2,
  bWriteData        = TRUE,
  quiet             = F,
  bWriteFIX         = TRUE,
  sortFREMDataset  = c("ID","TIME","FREMTYPE"),
  cstrKeepCols = c("ID","TIME","AMT","EVID","RATE","DV","FOOD","FREMTYPE"))

```

## Update initial estimates

```{r}
updateFREM(strFREMModel      = system.file("extdata/SimNeb/run31.mod", package = "PMXFrem"),
           numNonFREMThetas  = 7,
           numSkipOm         = 2,
           bWriteData        = FALSE,
           quiet             = F,
           strUpdateType     = "NoData")
```

