---
title: "Generate standard diagnostics for a FREM model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate standard diagnostics for a FREM model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages(library(GGally))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(PMXFrem))

theme_set(ggplot2::theme_bw(base_size=20))
```

## Introduction

Just like with any other model it is important to check that FREM models satisfy basic diagnostic tests, for example goodness-of-fit (GOF) plots of DV versus PRED and CWRES versus TIME. Unlike standard models, some of variables of interest are not directly available from the FREM model itself but will have to be derived based on an FFEM version of the model. 

## Create the FFEM model

Create the FFEM model for the set of covariates to include in the model to be evaluated.Consult the [companion vignette](create_FFEM_model.html) on how to set up the FFEM model and generate the necessary output.


```{r}
modDevDir <- system.file("extdata/SimNeb",package="PMXFrem")
dataFile  <- system.file("extdata/SimNeb/DAT-2-MI-PMX-2-onlyTYPE2-new.csv",package="PMXFrem")
fremRun   <- 31
baseRun   <- 30
fremFiles <- getFileNames(fremRun,modDevDir=modDevDir)
```

### Comparison of standard diagnostiocs input from Base, FREM and FFEM model

```{r}
xptabBase <- fread(paste0(modDevDir,"/xptab",baseRun),data.table=FALSE,check.names = T) %>%
  filter(BLQ==0,EVID==0) %>% 
    arrange(ID,TAD)

xptabFFEM <- fread(paste0(modDevDir,"/xptabmax0",fremRun),data.table=FALSE,check.names = T) %>%
  filter(BLQ==0,EVID==0) %>% 
  arrange(ID,TAD)

xptabFREM <- fread(paste0(modDevDir,"/xptab",fremRun),data.table=FALSE,check.names = T) %>%
  filter(BLQ==0,EVID==0,FREMTYPE==0) %>% 
    arrange(ID,TAD)

cwresData <- bind_cols(Base=xptabBase$CWRES, FREM=xptabFREM$CWRES, FFEM=xptabFFEM$CWRES) 
iwresData <- bind_cols(Base=xptabBase$CIWRES,FREM=xptabFREM$CIWRES,FFEM=xptabFFEM$CIWRES)
predData  <- bind_cols(Base=xptabBase$CPRED, FREM=xptabFREM$CPRED, FFEM=xptabFFEM$CPRED)
ipredData <- bind_cols(Base=xptabBase$CIPREDI,FREM=xptabFREM$CIPREDI,FFEM=xptabFFEM$CIPREDI)
```

It is only the population predictions that are the same between the three models. IPRED, CWRES, and IWRES are different
```{r,fig.width=10,fig.height=6}
ggpairs(predData)   + ggtitle("PRED")
```

The individual predictions are not the same.
```{r,fig.width=10,fig.height=6}
ggpairs(ipredData)  + ggtitle("IPRED")
```

The CWRES are not identical
```{r,fig.width=10,fig.height=6}
ggpairs(cwresData)  + ggtitle("CWRES")
```

The IWRES are not identical
```{r,fig.width=10,fig.height=6}

ggpairs(iwresData)  + ggtitle("IWRES")
```

## Creating basic GOF plots

```{r,fig.width=10,fig.height=10}

## DV vs PRED
p1 <- xptabFFEM %>% 
  ggplot(aes(exp(CPRED),exp(DV))) +
  geom_point(color="#508791") +
  geom_abline(slope = 1,intercept = 0,size=2) +
  geom_smooth(se = FALSE,color="red") +
  xlab("Populaiton predicitions") +
  ylab("Observations")


## DV vs IPRED
p2 <-xptabFFEM %>% 
  ggplot(aes(exp(CIPREDI),exp(DV))) +
  geom_point(color="#508791") +
  geom_abline(slope = 1,intercept = 0,size=2) +
  geom_smooth(se = FALSE,color="red") +
  xlab("Individual predicitions") +
  ylab("Observations")

## abs(IWRES) vs IPRED
p3 <- xptabFFEM %>% 
  ggplot(aes(abs(CIWRES),exp(CIPREDI))) +
  geom_point(color="#508791") +
  geom_smooth(se = FALSE,color="red") +
  xlab("Individual predicitions") +
  ylab("Absolute IWRES")

## CWRES vs IPRED
p4 <- xptabFFEM %>% 
  ggplot(aes(exp(CPRED),CWRES)) +
  geom_point(color="#508791") +
  geom_abline(slope = 0,intercept = 0,size=2) +
  geom_smooth(se = FALSE,color="red") +
  xlab("Population predicitions") +
  ylab("Conditional weighted residuals")

ggarrange(p1,p2,p3,p4,ncol=2,nrow=2)

```

## Creating ETA plots

Similarly to the standar diagnostics variables (e.g. CWRES) ETAs will also have to be adjusted to reflect the impact of covariates, i.e. the "ETA prims" needs to be computed. This is done by the `calcEtas` function. This function requires an "FFEM data object" from the `createFFEMmodel` function as input.

```{r}

data <- read_csv(dataFile,show_col_types = FALSE) %>%
    filter(BLQ!=1)

  ## Check with specified parameter names
FFEMdata <- createFFEMdata(runno            = fremRun,
                           modDevDir        = modDevDir,
                           parNames         = c("CL","V","MAT"),
                           numNonFREMThetas = 7,
                           numSkipOm        = 2,
                           dataFile         = data,
                           newDataFile      = NULL,
                           quiet            = TRUE)
```

Collect the ETAs and compute the ETA_proms
```{r}
ind_params <- calcEtas(runno            = fremRun,
                       modDevDir        = modDevDir,
                       numSkipOm        = 2,
                       numNonFREMThetas = 7,
                       FFEMData         = FFEMdata) %>% 
  arrange(ID)

print(head(ind_params))
```

Compare the ETAs and the ETA_prims
```{r,fig.width=10,fig.height=6}
ind_params %>% select(ETA3,ETA3_PRIM) %>% ggpairs()   + ggtitle("ETA CL")
```

```{r,fig.width=10,fig.height=6}
ind_params %>% select(ETA4,ETA4_PRIM) %>% ggpairs()   + ggtitle("ETA V")
```

```{r,fig.width=10,fig.height=6}
ind_params %>% select(ETA4,ETA5_PRIM) %>% ggpairs()   + ggtitle("ETA MAT")
```

All predicted covariate values will be the same as the observed values, provided there are informative observations for the subject in question (i.e. not only dosing records in PK). 
```{r fig.width=10,fig.height=6}

covariates <- data %>% 
  group_by(ID) %>% 
  mutate(haveObs = ifelse(any(DV>0),1,0)) %>% 
  ungroup %>% 
  distinct(ID,.keep_all = T) %>% 
  select(ID,getCovNames(fremFiles$mod)$orgCovNames,haveObs) %>% 
  arrange(ID)

tmp <- bind_cols(ind_params %>% select(ID,ETA3,ETA3_PRIM,WT),obsWT=covariates$WT,haveObs=covariates$haveObs) %>% 
  mutate(obsWT = ifelse(obsWT==-99,NA,obsWT)) %>% 
  mutate(missWT = ifelse(is.na(obsWT),"Missing WT","Not missing WT")) %>% 
  filter(haveObs==1) # Remove subects with inly dosing records

tmp %>% ggplot(aes(WT,obsWT,color=missWT)) +
  geom_point(show.legend = FALSE) +
  xlab("Predicted WT") +
  ylab("Observed WT")
```



Plotting the regular ETA from the FREM model versus WT gives a very different picture compared to plotting ETA prim.

```{r,fig.width=10,fig.height=6}

p1 <- tmp %>% 
  ggplot(aes(WT,ETA3)) +
  geom_point(aes(color=missWT)) +
  geom_smooth(se = FALSE,color="red") +
  ylab("ETA3 (CL)") +
  labs(color=NULL,linetype=NULL) + theme(legend.position="top") +
  scale_colour_manual(values = c("Missing WT"="orange","Not missing WT"="#508791")) 

p2 <- tmp %>% 
  ggplot(aes(WT,ETA3_PRIM)) +
  geom_point(aes(color=missWT)) +
  geom_smooth(se = FALSE,color="red") +
  ylab("ETA3 prim (CL)") +
  labs(color=NULL,linetype=NULL) + theme(legend.position="top") +
  scale_colour_manual(values = c("Missing WT"="orange","Not missing WT"="#508791")) 

ggarrange(p1,p2,ncol=2,nrow=1,common.legend = T)
```

```{r fig.width=10,fig.height=6}

tmp %>% mutate(missWT = factor(missWT,levels=c("Not missing WT","Missing WT"))) %>% 
  ggplot(aes(WT,fill=missWT)) +
  geom_histogram(position ="identity") +
  labs(fill=NULL,linetype=NULL) + theme(legend.position="top") +
  scale_fill_manual(values = c("Missing WT"="orange","Not missing WT"="#508791")) 
```

